### 2024-03-11 paper/paper.qmd
Initial modeling

---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - First author
  - Another author
thanks: "Code and data are available at: LINK."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
library(readxl)
library(here)

deaths_data <- read.csv(here("data/analysis_data/cleaned_file.csv"))
air_quality_data <- read.csv(here("data/analysis_data/cleaned_air_data.csv"))
```

# Introduction

You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.

The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it.

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false
# 
# modelsummary::modelsummary(
#   list(
#     "First model" = first_model
#   ),
#   statistic = "mad",
#   fmt = 2
# )
# Load the required library
library(MASS)
library(ggplot2)
library(dplyr)
library(cowplot)

# Check the column names in the merged data
colnames(merged_data)

# Make sure the column name for Total Deaths matches
# Adjust the column name accordingly if needed

# Create a negative binomial model
model_nb <- glm.nb(`Total.Deaths` ~ `OriginalValue` + `Cause`, data = merged_data)

# View the summary of the negative binomial model
summary(model_nb)

# Load the required library
library(ggplot2)

# Predict the values using the negative binomial model
merged_data$predicted_deaths <- predict(model_nb, type = "response")

# Plot the relationship between Air Quality and Predicted Deaths
ggplot(merged_data, aes(x = OriginalValue, y = predicted_deaths, color = Cause)) +
  geom_point() +
  geom_smooth(method = "glm.nb", formula = y ~ x, se = FALSE) +
  labs(x = "Air Quality Health Index", y = "Predicted Deaths", color = "Cause") +
  theme_minimal()

# Assuming you have loaded your data into 'air_quality_data' and 'deaths_data'

# Merge the air quality and deaths data by year
merged_data <- merge(air_quality_data, deaths_data, by = "Year")

# Plot the regression graph
regression_plot <- ggplot(merged_data, aes(x = OriginalValue, y = Total.Deaths)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +  # Add linear regression line
  labs(title = "Regression Graph: Air Quality vs Deaths",
       x = "Air Quality Index",
       y = "Total Deaths") +
  theme_minimal()

# Show the plot
print(regression_plot)

library(ggplot2)
library(tidyr)

# Your dataset
data <- data.frame(
  "Year" = c(2000:2021),
  "Category" = c(17.6, 17.5, 17.7, 19.9, 17.6, 12.9, 14.9, 14, 14.3, 16.1, 28, 26.9, 21.3, 18.1, 25.7, 31.9, 31.2, 25.8, 52, 21.7, 15.8, 32.9),
  "Provincial Average" = c(13.5, 9.5, 11.5, 11.5, 13.6, 8.9, 11.7, 10.3, 9.8, 11.9, 16.3, 11.3, 13.3, 10.3, 16.7, 17, 11.9, 14.9, 32.8, 15.3, 10.8, 26.2),
  "10th Percentile" = c(23.7, 22.4, 24.5, 33.5, 21.6, 16.6, 18.4, 16.8, 19.9, 22.4, 41.4, 60.5, 31.5, 24, 34.1, 48.1, 49.4, 34.9, 65.9, 28, 20.1, 42.4),
  "Calgary" = c(22, 20.8, 18.8, 35.5, 17.6, 13.2, 18.3, 16.2, 14.5, 19.5, 30.8, 24, 21.9, 20.8, 21.2, 28.1, 14.7, 34.6, 58.1, 19.2, 18.9, 44.2),
  "Edmonton" = c(24.9, 24, 23.4, 24, 21.4, 14.4, 17.8, 14.8, 18.6, 21.5, 46.4, 26.9, 22.3, 29.5, 26, 23.7, 22.7, 29, 49.1, 26.2, 20.6, 38.4),
  "Fort McMurray" = c(13.7, 15.9, 15.2, 11.5, 1.5, 12.7, 14.1, 15.1, 14.9, 13.9, 16.4, 72.4, 20.8, 17, 35.1, 40.9, 208.4, 15.9, 35.4, 21.5, 12.6, 25.9),
  "Grande Prairie" = c(NA, NA, NA, NA, 14.2, 14.8, 16.4, 11.9, 11.5, 14.7, 29.1, 20.3, 19.5, 17.6, 35.9, 20.9, 21, 26.3, 60.5, 23.8, 17.9, 29.7),
  "Lethbridge" = c(NA, NA, NA, NA, 15.7, 10.1, 12.3, 12, 9.8, NA, 25.7, 18.5, 23.4, NA, 16.5, 42, 10.7, 44.9, 58.4, 20.8, 20.5, 38.2),
  "Medicine Hat" = c(NA, NA, NA, NA, 13.8, 8.8, 9.5, 10.5, 10, 24.3, NA, 18, 22.3, 34.5, 20.4, 24.5, 16.5, 29.3, 43, 13.7, 16.3, 32.8),
  "Red Deer" = c(27, 20.9, 15.3, 20.9, 27, 27, 22.3, 11.8, 11.6, 27, 37.5, 34.1, 27, 27, 27, 27, 27, 27, 27, 27, 27, 27),
  "CAAQS" = c(27, 27, 27, 27, 27, 27, 27, 27, 27, 27, 27, 27, 27, 27, 27, 27, 27, 27, 27, 27, 27, 27)
)

# Reshape the data from wide to long format for plotting
data_long <- pivot_longer(data, cols = -Year, names_to = "Variable", values_to = "Value")

# Plot the average air index and total death rates over the years
graph <- ggplot(data_long, aes(x = Year, y = Value, color = Variable)) +
  geom_line() +
  labs(title = "Average Air Index and Total Death Rates Over the Years",
       x = "Year",
       y = "Value") +
  theme_minimal() +
  scale_color_manual(values = c("red", "blue", "green", "orange", "purple", "black", "brown", "lightblue","pink", "darkgreen", "lightgreen")) # Adjust color palette as needed

# Show the plot
print(graph)
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

\newpage

# References

### 2024-03-11 paper/paper.qmd
Made the model

@@
 
-analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
-
-analysis_data |> 
-  ggplot(aes(x = width, y = length)) +
-  geom_point(alpha = 0.8) +
-  theme_minimal() +
-  labs(x = "Wing width (mm)",
-       y = "Wing length (mm)")
+# analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
+# 
+# analysis_data |> 
+#   ggplot(aes(x = width, y = length)) +
+#   geom_point(alpha = 0.8) +
+#   theme_minimal() +
+#   labs(x = "Wing width (mm)",
+#        y = "Wing length (mm)")
 ```
@@
   readRDS(file = here::here("models/first_model.rds"))
+
+poisson_model <-
+  readRDS(file = here::here("models/poisson_model.rds"))
+
+causes_model <-
+  readRDS(file = here::here("models/causes_model.rds"))
 ```
@@
 #| warning: false
-# 
-# modelsummary::modelsummary(
-#   list(
-#     "First model" = first_model
-#   ),
-#   statistic = "mad",
-#   fmt = 2
-# )
-# Load the required library
-library(MASS)
-library(ggplot2)
-library(dplyr)
-library(cowplot)
 
-# Check the column names in the merged data
-colnames(merged_data)
-
-# Make sure the column name for Total Deaths matches
-# Adjust the column name accordingly if needed
-
-# Create a negative binomial model
-model_nb <- glm.nb(`Total.Deaths` ~ `OriginalValue` + `Cause`, data = merged_data)
-
-# View the summary of the negative binomial model
-summary(model_nb)
-
-# Load the required library
-library(ggplot2)
-
-# Predict the values using the negative binomial model
-merged_data$predicted_deaths <- predict(model_nb, type = "response")
-
-# Plot the relationship between Air Quality and Predicted Deaths
-ggplot(merged_data, aes(x = OriginalValue, y = predicted_deaths, color = Cause)) +
-  geom_point() +
-  geom_smooth(method = "glm.nb", formula = y ~ x, se = FALSE) +
-  labs(x = "Air Quality Health Index", y = "Predicted Deaths", color = "Cause") +
-  theme_minimal()
-
-# Assuming you have loaded your data into 'air_quality_data' and 'deaths_data'
-
-# Merge the air quality and deaths data by year
-merged_data <- merge(air_quality_data, deaths_data, by = "Year")
-
-# Plot the regression graph
-regression_plot <- ggplot(merged_data, aes(x = OriginalValue, y = Total.Deaths)) +
-  geom_point() +
-  geom_smooth(method = "lm", se = FALSE) +  # Add linear regression line
-  labs(title = "Regression Graph: Air Quality vs Deaths",
-       x = "Air Quality Index",
-       y = "Total Deaths") +
-  theme_minimal()
-
-# Show the plot
-print(regression_plot)
-
-library(ggplot2)
-library(tidyr)
-
-# Your dataset
-data <- data.frame(
-  "Year" = c(2000:2021),
-  "Category" = c(17.6, 17.5, 17.7, 19.9, 17.6, 12.9, 14.9, 14, 14.3, 16.1, 28, 26.9, 21.3, 18.1, 25.7, 31.9, 31.2, 25.8, 52, 21.7, 15.8, 32.9),
-  "Provincial Average" = c(13.5, 9.5, 11.5, 11.5, 13.6, 8.9, 11.7, 10.3, 9.8, 11.9, 16.3, 11.3, 13.3, 10.3, 16.7, 17, 11.9, 14.9, 32.8, 15.3, 10.8, 26.2),
-  "10th Percentile" = c(23.7, 22.4, 24.5, 33.5, 21.6, 16.6, 18.4, 16.8, 19.9, 22.4, 41.4, 60.5, 31.5, 24, 34.1, 48.1, 49.4, 34.9, 65.9, 28, 20.1, 42.4),
-  "Calgary" = c(22, 20.8, 18.8, 35.5, 17.6, 13.2, 18.3, 16.2, 14.5, 19.5, 30.8, 24, 21.9, 20.8, 21.2, 28.1, 14.7, 34.6, 58.1, 19.2, 18.9, 44.2),
-  "Edmonton" = c(24.9, 24, 23.4, 24, 21.4, 14.4, 17.8, 14.8, 18.6, 21.5, 46.4, 26.9, 22.3, 29.5, 26, 23.7, 22.7, 29, 49.1, 26.2, 20.6, 38.4),
-  "Fort McMurray" = c(13.7, 15.9, 15.2, 11.5, 1.5, 12.7, 14.1, 15.1, 14.9, 13.9, 16.4, 72.4, 20.8, 17, 35.1, 40.9, 208.4, 15.9, 35.4, 21.5, 12.6, 25.9),
-  "Grande Prairie" = c(NA, NA, NA, NA, 14.2, 14.8, 16.4, 11.9, 11.5, 14.7, 29.1, 20.3, 19.5, 17.6, 35.9, 20.9, 21, 26.3, 60.5, 23.8, 17.9, 29.7),
-  "Lethbridge" = c(NA, NA, NA, NA, 15.7, 10.1, 12.3, 12, 9.8, NA, 25.7, 18.5, 23.4, NA, 16.5, 42, 10.7, 44.9, 58.4, 20.8, 20.5, 38.2),
-  "Medicine Hat" = c(NA, NA, NA, NA, 13.8, 8.8, 9.5, 10.5, 10, 24.3, NA, 18, 22.3, 34.5, 20.4, 24.5, 16.5, 29.3, 43, 13.7, 16.3, 32.8),
-  "Red Deer" = c(27, 20.9, 15.3, 20.9, 27, 27, 22.3, 11.8, 11.6, 27, 37.5, 34.1, 27, 27, 27, 27, 27, 27, 27, 27, 27, 27),
-  "CAAQS" = c(27, 27, 27, 27, 27, 27, 27, 27, 27, 27, 27, 27, 27, 27, 27, 27, 27, 27, 27, 27, 27, 27)
+modelsummary::modelsummary(
+  list(
+    "First model" = first_model
+  ),
+  #statistic = "mad",
+  fmt = 2
 )
 
-# Reshape the data from wide to long format for plotting
-data_long <- pivot_longer(data, cols = -Year, names_to = "Variable", values_to = "Value")
+modelsummary::modelsummary(
+  list(
+    "First model" = poisson_model
+  ),
+  #statistic = "mad",
+  fmt = 2
+)
 
-# Plot the average air index and total death rates over the years
-graph <- ggplot(data_long, aes(x = Year, y = Value, color = Variable)) +
-  geom_line() +
-  labs(title = "Average Air Index and Total Death Rates Over the Years",
-       x = "Year",
-       y = "Value") +
-  theme_minimal() +
-  scale_color_manual(values = c("red", "blue", "green", "orange", "purple", "black", "brown", "lightblue","pink", "darkgreen", "lightgreen")) # Adjust color palette as needed
-
-# Show the plot
-print(graph)
+modelsummary::modelsummary(
+  list(
+    "First model" = causes_model
+  ),
+  #statistic = "mad",
+  fmt = 2
+)
 ```
@@
 
-pp_check(first_model) +
+# Create the posterior predictive check plot
+pp_plot <- pp_check(first_model) +
   theme_classic() +
@@
 
-posterior_vs_prior(first_model) +
+# Create the posterior vs prior plot
+posterior_plot <- posterior_vs_prior(first_model) +
   theme_minimal() +

### 2024-03-11 paper/paper.qmd
extra data removed, and new data of charts added

@@
 library(palmerpenguins)
-library(readxl)
 library(here)
+library(ggplot2)
 
+#Reading the data
 deaths_data <- read.csv(here("data/analysis_data/cleaned_file.csv"))
 air_quality_data <- read.csv(here("data/analysis_data/cleaned_air_data.csv"))
+avg_PM_conc_data <- read.csv(here("data/raw_data/avg PM2.5 conc data.csv"))
+peak_PM_conc_data <- read.csv(here("data/raw_data/peak PM2.5 conc data.csv"))
+avg_PM_conc_data <- avg_PM_conc_data %>% rename(Year = Category)
+peak_PM_conc_data <- peak_PM_conc_data %>% rename(Year = Category)
+
+# Assuming 'cleaned_air_data' is your air quality dataset and 'mortality_data' is your dataset of deaths.
+
+library(tidyverse)
+library(ggplot2)
+
+# Ensure you have the correct column names by checking the names of your data frames
+colnames(air_quality_data)
+colnames(deaths_data)
+
+# Merge the datasets
+combined_data <- left_join(deaths_data, air_quality_data, by = "year")
+
+head(combined_data)
+
+# 2. Mortality Trend Plot for a single cause
+# Repeat this plot changing the 'Cause' filter for each cause you are interested in.
+ggplot(combined_data %>% filter(cause == 'All other forms of chronic ischemic heart disease'), 
+       aes(x = year, y = total_deaths)) +  # Note the use of single quotes and backticks
+  geom_line() +
+  labs(title = "Trend in Deaths Due to Chronic Ischemic Heart Disease",
+       x = "Year",
+       y = "Total Deaths") +
+  theme_minimal()
+
+# 3. Relationship between Air Quality and Mortality Rates
+# This could be a scatter plot with a line representing the trend.
+ggplot(combined_data, aes(x = original_value, y = total_deaths, color = cause)) +  # Make sure the column name matches
+  geom_point() +
+  geom_smooth(method = "lm") +  # Add a linear trend line
+  facet_wrap(~cause) +  # Create separate plots for each cause
+  labs(title = "Relationship between Air Quality and Mortality Rates",
+       x = "Air Quality Health Index",
+       y = "Total Deaths") +
+  theme_minimal()
 ```
@@
 
-```{r}
-#| label: fig-bills
-#| fig-cap: Bills of penguins
-#| echo: false
-
-ggplot(penguins, aes(x = island, fill = species)) +
-  geom_bar(alpha = 0.8) +
-  scale_fill_manual(values = c("darkorange","purple","cyan4"),
-                    guide = "none") +
-  theme_minimal() +
-  facet_wrap(~species, ncol = 1) +
-  coord_flip()
-```
-
 Talk more about it.
@@
 (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
-
-```{r}
-#| label: fig-planes
-#| fig-cap: Relationship between wing length and width
-#| echo: false
-#| warning: false
-#| message: false
-
-# analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
-# 
-# analysis_data |> 
-#   ggplot(aes(x = width, y = length)) +
-#   geom_point(alpha = 0.8) +
-#   theme_minimal() +
-#   labs(x = "Wing width (mm)",
-#        y = "Wing length (mm)")
-```
 
@@
   readRDS(file = here::here("models/first_model.rds"))
-
-poisson_model <-
-  readRDS(file = here::here("models/poisson_model.rds"))
-
-causes_model <-
-  readRDS(file = here::here("models/causes_model.rds"))
 ```
@@
   ),
-  #statistic = "mad",
-  fmt = 2
-)
-
-modelsummary::modelsummary(
-  list(
-    "First model" = poisson_model
-  ),
-  #statistic = "mad",
-  fmt = 2
-)
-
-modelsummary::modelsummary(
-  list(
-    "First model" = causes_model
-  ),
-  #statistic = "mad",
+  statistic = "mad",
   fmt = 2
@@
 
-```{r}
-#| eval: true
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-ppcheckandposteriorvsprior
-#| layout-ncol: 2
-#| fig-cap: "Examining how the model fits, and is affected by, the data"
-#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
-
-# Create the posterior predictive check plot
-pp_plot <- pp_check(first_model) +
-  theme_classic() +
-  theme(legend.position = "bottom")
-
-# Create the posterior vs prior plot
-posterior_plot <- posterior_vs_prior(first_model) +
-  theme_minimal() +
-  scale_color_brewer(palette = "Set1") +
-  theme(legend.position = "bottom") +
-  coord_flip()
-```
-
 ## Diagnostics

### 2024-03-11 scripts/02-data_cleaning.R

@@ -27,6 +27,9 @@ colnames(cleaned_df) <- c("Year", "Cause", "Ranking", "Total Deaths")
 cleaned_df <- cleaned_df %>%
   filter(`Year` >= 2012)
 
+#cleaning names 
+cleaned_df = clean_names(cleaned_df)
+
 # Write the cleaned data to a new CSV file
 write_csv(cleaned_df, "data/analysis_data/cleaned_file.csv")
 
@@ -46,6 +49,11 @@ cleaned_data <- cleaned_data %>%
   arrange(Year)
 cleaned_data <- cleaned_data %>%
   filter(`Year` >= 2012)
+
+#clean names
+
+cleaned_data = clean_names(cleaned_data)
+
 
 # Save the cleaned dataset to a CSV file
 write.csv(cleaned_data, file = "data/analysis_data/cleaned_air_data.csv", row.names = FALSE)

### 2024-03-11 scripts/04-model.R
fake commit

@@ -13,25 +13,61 @@ library(tidyverse)
 library(rstanarm)
 
 #### Read data ####
-analysis_data <- read_csv("data/analysis_data/analysis_data.csv")
+deaths_data <- read.csv(here("data/analysis_data/cleaned_file.csv"))
+air_quality_data <- read.csv(here("data/analysis_data/cleaned_air_data.csv"))
+
 
 ### Model data ####
-first_model <-
-  stan_glm(
-    formula = flying_time ~ length + width,
-    data = analysis_data,
-    family = gaussian(),
-    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_aux = exponential(rate = 1, autoscale = TRUE),
-    seed = 853
-  )
+# first_model <-
+#   stan_glm(
+#     formula = flying_time ~ length + width,
+#     data = analysis_data,
+#     family = gaussian(),
+#     prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
+#     prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
+#     prior_aux = exponential(rate = 1, autoscale = TRUE),
+#     seed = 853
+#   )
 
+# 
+# merged_data <- merge(air_quality_data, deaths_data, by = "year")  # Adjust the common key as needed
+# 
+# print(merged_data)
+# # Fit negative binomial regression model
+# total_deaths_model_neg_binomial <- stan_glm(
+#   total_deaths ~ original_value + cause,
+#   data = merged_data,
+#   family = neg_binomial_2(link = "log"),
+#   seed = 123  # Set your desired seed value
+# )
+library(tidyverse)
+library(readr)
+library(MASS)
+
+# Convert year and air_quality_health_index to numeric
+data <- air_quality_data %>%
+  mutate(year = as.numeric(year),
+         air_quality_health_index = as.numeric(air_quality_health_index))
+
+# Fit negative binomial regression model
+first_model <- glm.nb(air_quality_health_index ~ year, data = data)
+
+
+# Fit Poisson regression model
+poisson_model <- glm(air_quality_health_index ~ year, data = data, family = poisson())
+
+# For the four causes and total deaths
+# Assuming the dataset for causes and total deaths is named 'causes_data'
+
+causes_model <- glm.nb(total_deaths ~ year + cause, data = deaths_data)
 
[…]

### 2024-03-13 paper/paper.qmd
Model new connections and relations

--- 2024-03-11 paper/paper.qmd
@@
 air_quality_data <- read.csv(here("data/analysis_data/cleaned_air_data.csv"))
+data <- read.csv(here("data/analysis_data/cleaned_chart_data.csv"))
+data_peak <- read.csv(here("data/analysis_data/cleaned_peak_data.csv"))
 ```
@@
 
-ggplot(penguins, aes(x = island, fill = species)) +
-  geom_bar(alpha = 0.8) +
-  scale_fill_manual(values = c("darkorange","purple","cyan4"),
-                    guide = "none") +
-  theme_minimal() +
-  facet_wrap(~species, ncol = 1) +
-  coord_flip()
+# Plot the regression graph
+ggplot(data, aes(x = total_deaths, y = provincial_average)) +
+  geom_point() +
+  geom_smooth(method = "lm", se = FALSE) +  # Add linear regression line without confidence interval
+  labs(x = "Total Deaths", y = "Provincial Average", title = "Regression Graph") +
+  theme_minimal()
+
+# Plot the regression graph
+ggplot(data_peak, aes(x = total_deaths, y = provincial_average)) +
+  geom_point() +
+  geom_smooth(method = "lm", se = FALSE) +  # Add linear regression line without confidence interval
+  labs(x = "Total Deaths", y = "Provincial Average", title = "Regression Graph") +
+  theme_minimal()
+
+# Plot the regression graph
+ggplot(data_peak, aes(x = provincial_average, y = total_deaths)) +
+  geom_point() +
+  geom_smooth(method = "lm", se = FALSE) +  # Add linear regression line without confidence interval
+  labs(x = "Provincial Average", y = "Total Deaths", title = "Regression Graph") +
+  theme_minimal()
+
+# Plot the regression graph
+ggplot(data, aes(x = provincial_average, y = total_deaths)) +
+  geom_point() +
+  geom_smooth(method = "lm", se = FALSE) +  # Add linear regression line without confidence interval
+  labs(x = "Provincial Average", y = "Total Deaths", title = "Regression Graph") +
+  theme_minimal()
+
 ```
@@
   readRDS(file = here::here("models/causes_model.rds"))
+
+p_average_model <-
+  readRDS(file = here::here("models/p_average_model.rds"))
 ```
@@
     "First model" = causes_model
+  ),
+  #statistic = "mad",
+  fmt = 2
+)
+
+modelsummary::modelsummary(
+  list(
+    "First model" = p_average_model
   ),
@@
 
-# Create the posterior predictive check plot
-pp_plot <- pp_check(first_model) +
-  theme_classic() +
+pp_check(p_average_model) +
   theme(legend.position = "bottom")
 
-# Create the posterior vs prior plot
-posterior_plot <- posterior_vs_prior(first_model) +
-  theme_minimal() +
+# Create the posterior predictive check plot with the specified theme
+pp_plot <- pp_check(p_average_model, theme = theme_classic()) +
+  theme(legend.position = "bottom")
+
+# Create the posterior vs prior plot with the specified theme
+posterior_plot <- posterior_vs_prior(p_average_model, theme = theme_minimal()) +
   scale_color_brewer(palette = "Set1") +

### 2024-03-13 scripts/02-data_cleaning.R

@@ -14,7 +14,7 @@ library(dplyr)
 #### Clean data ####
 # first dataset
 df <- read_csv("data/raw_data/deaths-leading-causes.csv", skip = 2)
-
+df1 <- read.csv("data/raw_data/deaths-leading-causes.csv", skip = 2)
 # Filter the data to include only rows where the Cause is "Acute myocardial infarction" or "Malignant neoplasms of trachea, bronchus and lung"
 filtered_df <- df[df$Cause %in% c("Acute myocardial infarction", "Malignant neoplasms of trachea, bronchus and lung", "Other chronic obstructive pulmonary disease", "All other forms of chronic ischemic heart disease"), ]
 
@@ -52,3 +52,54 @@ cleaned_data = clean_names(cleaned_data)
 
 # Save the cleaned dataset to a CSV file
 write.csv(cleaned_data, file = "data/analysis_data/cleaned_air_data.csv", row.names = FALSE)
+
+
+## extra dataset ##
+
+
+
+# Filter the data to include only rows where the Cause is "Acute myocardial infarction" or "Malignant neoplasms of trachea, bronchus and lung"
+filtered_df <- df[df$Cause %in% c("Malignant neoplasms of trachea, bronchus and lung"), ]
+
+# Select only the necessary columns: Year, Cause, Ranking, and Total Deaths
+cleaned_df <- filtered_df[, c("Calendar Year", "Cause", "Ranking", "Total Deaths")]
+
+# Rename the columns
+colnames(cleaned_df) <- c("Year", "Cause", "Ranking", "Total Deaths")
+cleaned_df <- cleaned_df %>%
+  filter(`Year` < 2022)
+cleaned_df = clean_names(cleaned_df)
+
+chart <- read.csv("data/raw_data/chart.csv")
+
+names(chart)[1] <- "year"
+# Select only the necessary columns: Year, Cause, Ranking, and Total Deaths
+chart <- chart [, c("year", "Provincial.Average")]
+chart <- chart %>%
+  filter(`year` >= 2001)
+chart = clean_names(chart)
+
+# Merge datasets based on the year column
+merged_data <- merge(cleaned_df, chart, by.x = "year", by.y = "year", all.x = TRUE)
+
+# Save the cleaned dataset to a CSV file
+write.csv(merged_data, file = "data/analysis_data/cleaned_chart_data.csv", row.names = FALSE)
+
+
+## P2.5 PEAK GRAPHS
+
[…]

### 2024-03-13 scripts/04-model.R

@@ -15,7 +15,8 @@ library(rstanarm)
 #### Read data ####
 deaths_data <- read.csv(here("data/analysis_data/cleaned_file.csv"))
 air_quality_data <- read.csv(here("data/analysis_data/cleaned_air_data.csv"))
-
+data_average <- read.csv(here("data/analysis_data/cleaned_chart_data.csv"))
+data_peak <- read.csv(here("data/analysis_data/cleaned_peak_data.csv"))
 
 ### Model data ####
 # first_model <-
@@ -29,17 +30,7 @@ air_quality_data <- read.csv(here("data/analysis_data/cleaned_air_data.csv"))
 #     seed = 853
 #   )
 
-# 
-# merged_data <- merge(air_quality_data, deaths_data, by = "year")  # Adjust the common key as needed
-# 
-# print(merged_data)
-# # Fit negative binomial regression model
-# total_deaths_model_neg_binomial <- stan_glm(
-#   total_deaths ~ original_value + cause,
-#   data = merged_data,
-#   family = neg_binomial_2(link = "log"),
-#   seed = 123  # Set your desired seed value
-# )
+
 library(tidyverse)
 library(readr)
 library(MASS)
@@ -61,6 +52,24 @@ poisson_model <- glm(air_quality_health_index ~ year, data = data, family = pois
 
 causes_model <- glm.nb(total_deaths ~ year + cause, data = deaths_data)
 
+## model for concentrations
+
+
+# # Fit negative binomial regression model using stan_glm
+# negative_binomial_model <- stan_glm(total_deaths ~ provincial_average, 
+#                                     data = data, 
+#                                     family = neg_binomial_2(), 
+#                                     seed = 123)
+# 
+# # Summary of the model
+# summary(negative_binomial_model)
+
+# Fit negative binomial regression model
+negative_binomial_model <- glm.nb(total_deaths ~ provincial_average, data = data_average)
+
+# Summary of the model
+summary(negative_binomial_model)
+
 #### Save model ####
 saveRDS(
   first_model,
@@ -70,4 +79,7 @@ saveRDS(poisson_model,
         file = "models/poisson_model.rds" )
 saveRDS(causes_model,
         file = "models/causes_model.rds" )
+saveRDS(negative_binomial_model,
+        file = "models/p_average_model.rds" )
+
 

### 2024-03-14 paper/paper.qmd
More models and changes to the cleaning

@@
 #| warning: false
-
+library(modelsummary)
 modelsummary::modelsummary(
@@
 
+```{r}
+#| eval: true
+#| echo: false
+#| message: false
+#| warning: false
+#| label: fig-ppcheckandposteriorvsprior
+#| layout-ncol: 2
+#| fig-cap: "Examining how the model fits, and is affected by, the data"
+#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
+
+library(rstanarm)
+library(here)
+library(ggplot2)
+library(modelsummary)
+
+# Assuming you've already read your models as before
+neg_binom_model <- readRDS(file = here::here("models/neg_binom_model.rds"))
+stan_neg_binom_model <- readRDS(file = here::here("models/stan_neg_binom_model.rds"))
+stan_poisson_model <-readRDS(file = here::here("models/stan_poisson_model.rds"))
+first_model <-
+  readRDS(file = here::here("models/first_model.rds"))
+
+poisson_model <-
+  readRDS(file = here::here("models/poisson_model.rds"))
+
+if (!requireNamespace("broom.mixed", quietly = TRUE)) {
+  install.packages("broom.mixed")
+}
+
+library(broom.mixed)
+
+# Summarize the models
+modelsummary(
+  list(
+    "Poisson" = cause_of_death_alberta_poisson,
+    "Negative binomial" = cause_of_death_alberta_neg_binomial
+  ),
+  coef_map = coef_short_names,
+  estimate_function = function(model) {
+    broom.mixed::tidy(model, parameters = "all", conf.int = TRUE)
+  }
+)
+
+pp_check(first_model) +
+  theme(legend.position = "bottom")
+
+pp_check(poisson_model) +
+  theme(legend.position = "bottom")
+
+pp_stan_neg_binom_model <- pp_check(stan_neg_binom_model)
+pp_stan_neg_binom_model + theme(legend.position = "bottom")
+
+pp_stan_poisson_model <- pp_check(stan_poisson_model)
+pp_stan_poisson_model + theme(legend.position = "bottom")
+
+```
 # Discussion
@@
 
-pp_check(p_average_model) +
-  theme(legend.position = "bottom")
-
+# library(bayesplot)
+# 
+# ppc <- pp_check(p_average_model, type = "dens_overlay")
+# 
+# # Add a classic theme to the plot
+# #ppc + theme_classic()
+# 
+# # Print the plot
+# print(ppc)
 # Create the posterior predictive check plot with the specified theme

### 2024-03-14 scripts/02-data_cleaning.R

@@ -86,6 +86,8 @@ merged_data <- merge(cleaned_df, chart, by.x = "year", by.y = "year", all.x = TR
 write.csv(merged_data, file = "data/analysis_data/cleaned_chart_data.csv", row.names = FALSE)
 
 
+
+
 ## P2.5 PEAK GRAPHS
 
 chart <- read.csv("data/raw_data/ppeak.csv")
@@ -103,3 +105,78 @@ merged_data <- merge(cleaned_df, chart, by.x = "year", by.y = "year", all.x = TR
 # Save the cleaned dataset to a CSV file
 write.csv(merged_data, file = "data/analysis_data/cleaned_peak_data.csv", row.names = FALSE)
 
+
+## merge files for all the lung causes 
+
+# Filter the data to include only rows where the Cause is "Acute myocardial infarction" or "Malignant neoplasms of trachea, bronchus and lung"
+filtered_df <- df[df$Cause %in% c("Malignant neoplasms of trachea, bronchus and lung", "Other chronic obstructive pulmonary disease" ), ]
+
+filtered_df
+# Select only the necessary columns: Year, Cause, Ranking, and Total Deaths
+cleaned_df <- filtered_df[, c("Calendar Year", "Cause", "Ranking", "Total Deaths")]
+
+# Rename the columns
+colnames(cleaned_df) <- c("Year", "Cause", "Ranking", "Total Deaths")
+cleaned_df <- cleaned_df %>%
+  filter(`Year` < 2022)
+cleaned_df = clean_names(cleaned_df)
+
+# Group by year and sum the total deaths
+cleaned_df <- cleaned_df %>%
+  group_by(year) %>%
+  summarise(total_deaths = sum(total_deaths))
+
+chart <- read.csv("data/raw_data/chart.csv")
+
+names(chart)[1] <- "year"
+# Select only the necessary columns: Year, Cause, Ranking, and Total Deaths
+chart <- chart [, c("year", "Provincial.Average")]
+chart <- chart %>%
+  filter(`year` >= 2001)
+chart = clean_names(chart)
+
+# Merge datasets based on the year column
+merged_data <- merge(cleaned_df, chart, by.x = "year", by.y = "year", all.x = TRUE)
+
+# Save the cleaned dataset to a CSV file
+write.csv(merged_data, file = "data/analysis_data/merged_data.csv", row.names = FALSE)
+
+
+
+
+
+## merge files for all the heart causes 
+
[…]

### 2024-03-14 scripts/04-model.R

@@ -1,3 +1,93 @@
+# #### Preamble ####
+# # Purpose: Models... [...UPDATE THIS...]
+# # Author: Rohan Alexander [...UPDATE THIS...]
+# # Date: 11 February 2023 [...UPDATE THIS...]
+# # Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# # License: MIT
+# # Pre-requisites: [...UPDATE THIS...]
+# # Any other information needed? [...UPDATE THIS...]
+# 
+# 
+# #### Workspace setup ####
+# library(tidyverse)
+# library(rstanarm)
+# 
+# #### Read data ####
+# deaths_data <- read.csv(here("data/analysis_data/cleaned_file.csv"))
+# air_quality_data <- read.csv(here("data/analysis_data/cleaned_air_data.csv"))
+# data_average <- read.csv(here("data/analysis_data/cleaned_chart_data.csv"))
+# data_peak <- read.csv(here("data/analysis_data/cleaned_peak_data.csv"))
+# 
+# ### Model data ####
+# # first_model <-
+# #   stan_glm(
+# #     formula = flying_time ~ length + width,
+# #     data = analysis_data,
+# #     family = gaussian(),
+# #     prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
+# #     prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
+# #     prior_aux = exponential(rate = 1, autoscale = TRUE),
+# #     seed = 853
+# #   )
+# 
+# 
+# library(tidyverse)
+# library(readr)
+# library(MASS)
+# 
+# # Convert year and air_quality_health_index to numeric
+# data <- air_quality_data %>%
+#   mutate(year = as.numeric(year),
+#          air_quality_health_index = as.numeric(air_quality_health_index))
+# 
+# # Fit negative binomial regression model
+# first_model <- glm.nb(air_quality_health_index ~ year, data = data)
+# 
+# 
+# # Fit Poisson regression model
+# poisson_model <- glm(air_quality_health_index ~ year, data = data, family = poisson())
+# 
+# # For the four causes and total deaths
+# # Assuming the dataset for causes and total deaths is named 'causes_data'
+# 
+# causes_model <- glm.nb(total_deaths ~ year + cause, data = deaths_data)
+# 
+# ## model for concentrations
+# 
+# 
+# # # Fit negative binomial regression model using stan_glm
[…]

### 2024-03-14 paper/paper.qmd
revise

--- 2024-03-13 paper/paper.qmd
@@
 data_peak <- read.csv(here("data/analysis_data/cleaned_peak_data.csv"))
+data_lung <- read.csv(here("data/analysis_data/merged_data.csv"))
+data_heart <- read.csv(here("data/analysis_data/merged_heart_data.csv"))
 ```
@@
   labs(x = "Provincial Average", y = "Total Deaths", title = "Regression Graph") +
+  theme_minimal()
+
+# Plot the regression graph for all lungs causes
+ggplot(data_lung, aes(x = provincial_average, y = total_deaths)) +
+  geom_point() +
+  geom_smooth(method = "lm", se = FALSE) +  # Add linear regression line without confidence interval
+  labs(x = "Provincial Average", y = "Total Deaths", title = "Regression Lung") +
+  theme_minimal()
+
+# Plot the regression graph for all heart causes
+ggplot(data, aes(x = provincial_average, y = total_deaths)) +
+  geom_point() +
+  geom_smooth(method = "lm", se = FALSE) +  # Add linear regression line without confidence interval
+  labs(x = "Provincial Average", y = "Total Deaths", title = "Regression Heart") +
   theme_minimal()
@@
 #| message: false
+# 
+# library(rstanarm)
+# 
+# first_model <-
+#   readRDS(file = here::here("models/first_model.rds"))
+# 
+# poisson_model <-
+#   readRDS(file = here::here("models/poisson_model.rds"))
+# 
+# causes_model <-
+#   readRDS(file = here::here("models/causes_model.rds"))
+# 
+# p_average_model <-
+#   readRDS(file = here::here("models/p_average_model.rds"))
+#
 
 library(rstanarm)
+library(here)
+library(ggplot2)
+library(modelsummary)
 
+# Assuming you've already read your models as before
+neg_binom_model <- readRDS(file = here::here("models/neg_binom_model.rds"))
+stan_neg_binom_model <- readRDS(file = here::here("models/stan_neg_binom_model.rds"))
+stan_poisson_model <-readRDS(file = here::here("models/stan_poisson_model.rds"))
+peak_neg_binom_model <- readRDS(file = here::here("models/peak_neg_binom_model.rds"))
+peak_poisson_model <-readRDS(file = here::here("models/peak_poisson_model.rds"))
 first_model <-
@@
 
-causes_model <-
-  readRDS(file = here::here("models/causes_model.rds"))
-
-p_average_model <-
-  readRDS(file = here::here("models/p_average_model.rds"))
 ```
@@
 #| warning: false
+library(modelsummary)
+# modelsummary::modelsummary(
+#   list(
+#     "First model" = first_model
+#   ),
+#   #statistic = "mad",
+#   fmt = 2
+# )
+# 
+# modelsummary::modelsummary(
+#   list(
+#     "First model" = poisson_model
+#   ),
+#   #statistic = "mad",
+#   fmt = 2
+# )
+# 
+# modelsummary::modelsummary(
+#   list(
+#     "First model" = causes_model
+#   ),
+#   #statistic = "mad",
+#   fmt = 2
+# )
+# 
+# modelsummary::modelsummary(
+#   list(
+#     "First model" = p_average_model
+#   ),
+#   #statistic = "mad",
+#   fmt = 2
+# )
+#
 
-modelsummary::modelsummary(
-  list(
-    "First model" = first_model
-  ),
-  #statistic = "mad",
-  fmt = 2
+if (!requireNamespace("modelsummary", quietly = TRUE)) {
+  install.packages("modelsummary")
+}
+library(modelsummary)
+
+if (!requireNamespace("broom.mixed", quietly = TRUE)) {
+  install.packages("broom.mixed")
+}
+
+library(broom.mixed)
+
+coef_short_names <- 
+  c("causeAll other forms of chronic ischemic heart disease" = "Ischemic Heart Disease",
+    "causeMalignant neoplasms of trachea, bronchus and lung" = "Trachea/Bronchus/Lung Cancer",
+    "causeOrganic dementia" = "Organic Dementia",
+    "causeOther chronic obstructive pulmonary disease" = "COPD"
   )
 
-modelsummary::modelsummary(
+# Summarize the models
+modelsummary(
   list(
-    "First model" = poisson_model
+    "Poisson" = poisson_model,
+    "Negative binomial" = first_model
   ),
-  #statistic = "mad",
-  fmt = 2
+  coef_map = coef_short_names,
+  estimate_function = function(model) {
+    broom.mixed::tidy(model, parameters = "all", conf.int = TRUE)
+  }
 )
 
-modelsummary::modelsummary(
-  list(
-    "First model" = causes_model
-  ),
-  #statistic = "mad",
-  fmt = 2
-)
-
-modelsummary::modelsummary(
-  list(
-    "First model" = p_average_model
-  ),
-  #statistic = "mad",
-  fmt = 2
-)
 ```
-
 # Discussion
@@
 
-pp_check(p_average_model) +
+pp_check(first_model) +
   theme(legend.position = "bottom")
 
-# Create the posterior predictive check plot with the specified theme
-pp_plot <- pp_check(p_average_model, theme = theme_classic()) +
+pp_check(poisson_model) +
   theme(legend.position = "bottom")
 
-# Create the posterior vs prior plot with the specified theme
-posterior_plot <- posterior_vs_prior(p_average_model, theme = theme_minimal()) +
-  scale_color_brewer(palette = "Set1") +
-  theme(legend.position = "bottom") +
-  coord_flip()
+pp_stan_neg_binom_model <- pp_check(stan_neg_binom_model)
+pp_stan_neg_binom_model + theme(legend.position = "bottom")
+
+pp_stan_poisson_model <- pp_check(stan_poisson_model)
+pp_stan_poisson_model + theme(legend.position = "bottom")
+
+pp_check(peak_poisson_model) +
+  theme(legend.position = "bottom")
+
+pp_stan_neg_binom_model <- pp_check(peak_poisson_model)
+pp_stan_neg_binom_model + theme(legend.position = "bottom")
+
+pp_check(peak_neg_binom_model) +
+  theme(legend.position = "bottom")
+
+pp_stan_neg_binom_model <- pp_check(peak_neg_binom_model)
+pp_stan_neg_binom_model + theme(legend.position = "bottom")
+
 ```

### 2024-03-14 paper/paper.qmd
more updates

@@
 ---
-title: "My title"
+title: "Respiration Related Death Cause show Signficant Relation to Increasing "
 subtitle: "My subtitle if needed"
 author: 
-  - First author
-  - Another author
-thanks: "Code and data are available at: LINK."
+  - Vanshika Vanshika
+  - Shivank Goel
+  - Navya Hooda
+thanks: "Code and data are available at: https://github.com/shivankgoel003/Mortality-in-Alberta."
 date: today
@@
 # Plot the regression graph
-ggplot(data, aes(x = total_deaths, y = provincial_average)) +
-  geom_point() +
-  geom_smooth(method = "lm", se = FALSE) +  # Add linear regression line without confidence interval
-  labs(x = "Total Deaths", y = "Provincial Average", title = "Regression Graph") +
-  theme_minimal()
-
-# Plot the regression graph
-ggplot(data_peak, aes(x = total_deaths, y = provincial_average)) +
-  geom_point() +
-  geom_smooth(method = "lm", se = FALSE) +  # Add linear regression line without confidence interval
-  labs(x = "Total Deaths", y = "Provincial Average", title = "Regression Graph") +
-  theme_minimal()
-
-# Plot the regression graph
-ggplot(data_peak, aes(x = provincial_average, y = total_deaths)) +
+ggplot(data, aes(x = provincial_average, y = total_deaths)) +
   geom_point() +
@@
 
-# Plot the regression graph
-ggplot(data, aes(x = provincial_average, y = total_deaths)) +
-  geom_point() +
-  geom_smooth(method = "lm", se = FALSE) +  # Add linear regression line without confidence interval
-  labs(x = "Provincial Average", y = "Total Deaths", title = "Regression Graph") +
-  theme_minimal()
+```
+
+```{r}
+#| label: fig-bills
+#| fig-cap: Bills of penguins
+#| echo: false
 
@@
   theme_minimal()
+```
 
+```{r}
+#| label: fig-bills
+#| fig-cap: Bills of penguins
+#| echo: false
 # Plot the regression graph for all heart causes
@@
 
-# analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
-# 
-# analysis_data |> 
-#   ggplot(aes(x = width, y = length)) +
-#   geom_point(alpha = 0.8) +
-#   theme_minimal() +
-#   labs(x = "Wing width (mm)",
-#        y = "Wing length (mm)")
+# Assuming your data frame is named 'cleaned_avg_PM25_conc' and is already filtered and renamed.
+# Convert factors to character to avoid issues in ggplot
+data[] <- lapply(data, function(x) if(is.factor(x)) as.character(x) else x)
+# Creating the plot with ggplot
+p <- ggplot(data, aes(x = year)) +
+  geom_line(aes(y = provincial_average, colour = "Provincial Average")) +
+  geom_line(aes(y = x10th_percentile, colour = "10th Percentile")) +
+  geom_line(aes(y = x90th_percentile, colour = "90th Percentile")) +
+  labs(title = "Annual Trends of PM2.5 Concentrations in Alberta: Provincial Averages and Percentile Ranges (2012-2021)",
+       x = "Years", y = "PM2.5 Concentration (μg/m³)") +
+  scale_colour_manual(values = c("Provincial Average" = "blue", "10th Percentile" = "green", "90th Percentile" = "red")) +
+  theme_minimal()
+
+# Display the plot
+p
+
+library(MASS)
+
+air_quality_data <- air_quality_data[air_quality_data$air_quality_health_index != "10+",]
+air_quality_data$air_quality_health_index <- as.numeric(air_quality_data$air_quality_health_index)
+
+# Calculate the average Air Quality Health Index per year
+average_aqi_by_year <- air_quality_data %>%
+  group_by(year) %>%
+  summarize(average_aqi = mean(air_quality_health_index))
+
+combined_data <- merge(deaths_data, average_aqi_by_year, by = "year")
+
+# Plot scatter plot of mortality rates vs. air quality health index
+ggplot(combined_data, aes(x = average_aqi, y = total_deaths)) +
+  geom_point() +
+  geom_smooth(method = "lm") +
+  facet_wrap(~cause) +
+  labs(title = "Scatter Plot of Mortality Rates vs. Air Quality Health Index",
+       x = "Average Air Quality Health Index",
+       y = "Total Deaths")
+
+# Fit the model
+model <- glm.nb(total_deaths ~ average_aqi + factor(year) + factor(cause), data = combined_data)
+
+# Model Diagnostics
+par(mfrow=c(2,2))
+plot(model)
+
+# Add Predictions to the combined data
+# Note: predict() function might throw an error if there are new factor levels in the test set that were not present in the training set
+# To prevent this, the newdata argument should only include levels that were present in the model fitting
+combined_data$predicted_deaths <- predict(model, newdata = combined_data, type = "response")
+
+# Plot Observed vs. Predicted Deaths
+ggplot(combined_data, aes(x = total_deaths, y = predicted_deaths)) +
+  geom_point() +
+  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
+  labs(x = "Observed Deaths", y = "Predicted Deaths") +
+  theme_minimal()
+
+# Effect of AQHI on Predicted Deaths
+ggplot(combined_data, aes(x = average_aqi, y = predicted_deaths, color = factor(year))) +
+  geom_point() +
+  geom_line(aes(group = factor(year))) + # Adding lines to connect points by year
+  labs(x = "Average Air Quality Health Index", y = "Predicted Deaths") +
+  theme_minimal() +
+  facet_wrap(~cause)
+
+# Merge datasets on 'year'
+combined_data <- merge(deaths_data, average_aqi_by_year, by = "year")
+
+# Plot scatter plot of mortality rates vs. air quality health index
+ggplot(combined_data, aes(x = average_aqi, y = total_deaths)) +
+  geom_point() +
+  geom_smooth(method = "lm") +
+  facet_wrap(~cause) +
+  labs(title = "Scatter Plot of Mortality Rates vs. Air Quality Health Index",
+       x = "Average Air Quality Health Index",
+       y = "Total Deaths")
+
+# Summarize total deaths for each disease by year
+total_deaths_by_year <- deaths_data %>%
+  group_by(year, cause) %>%
+  summarize(total_deaths = sum(total_deaths))
+
+# Plot for mortality data
+ggplot(total_deaths_by_year, aes(x = as.factor(year), y = total_deaths, group = cause, color = cause)) +
+  geom_line() +
+  labs(title = "Total Deaths by Year for Each Disease",
+       x = "Year",
+       y = "Total Deaths")
+
+# Ensure the 'year' column is in numeric format for proper line plotting
+average_aqi_by_year$year <- as.numeric(as.character(average_aqi_by_year$year))
+
+# Plotting with ggplot2
+ggplot(average_aqi_by_year, aes(x = year, y = average_aqi)) +
+  geom_line() +  # for lines
+  geom_point() + # for points
+  labs(title = "Average Air Quality Health Index by Year",
+       x = "Year",
+       y = "Average Air Quality Index") +
+  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) # Optional: Adjust x-axis breaks
+
+ggplot(average_aqi_by_year, aes(x = as.factor(year), y = average_aqi, fill = as.factor(year))) +
+  geom_bar(stat = "identity") +
+  labs(title = "Average Air Quality Health Index by Year",
+       x = "Year",
+       y = "Average AQHI") +
+  theme_minimal()
+
 ```
@@
 #| message: false
-# 
-# library(rstanarm)
-# 
-# first_model <-
-#   readRDS(file = here::here("models/first_model.rds"))
-# 
-# poisson_model <-
-#   readRDS(file = here::here("models/poisson_model.rds"))
-# 
-# causes_model <-
-#   readRDS(file = here::here("models/causes_model.rds"))
-# 
-# p_average_model <-
-#   readRDS(file = here::here("models/p_average_model.rds"))
-#
 
@@
 stan_poisson_model <-readRDS(file = here::here("models/stan_poisson_model.rds"))
-peak_neg_binom_model <- readRDS(file = here::here("models/peak_neg_binom_model.rds"))
-peak_poisson_model <-readRDS(file = here::here("models/peak_poisson_model.rds"))
+
+heart_model <- readRDS(file = here::here("models/heart_model.rds"))
+lung_model <-readRDS(file = here::here("models/lung_model.rds"))
+
 first_model <-
@@
 #| label: tbl-modelresults
-#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
+#| tbl-cap: "Model of "
 #| warning: false
 library(modelsummary)
-# modelsummary::modelsummary(
-#   list(
-#     "First model" = first_model
-#   ),
-#   #statistic = "mad",
-#   fmt = 2
-# )
-# 
-# modelsummary::modelsummary(
-#   list(
-#     "First model" = poisson_model
-#   ),
-#   #statistic = "mad",
-#   fmt = 2
-# )
-# 
-# modelsummary::modelsummary(
-#   list(
-#     "First model" = causes_model
-#   ),
-#   #statistic = "mad",
-#   fmt = 2
-# )
-# 
-# modelsummary::modelsummary(
-#   list(
-#     "First model" = p_average_model
-#   ),
-#   #statistic = "mad",
-#   fmt = 2
-# )
-#
+
+modelsummary::modelsummary(
+  list(
+    "heart causes model" = heart_model
+  ),
+  #statistic = "mad",
+  fmt = 2
+)
+
+modelsummary::modelsummary(
+  list(
+    "Lung Causes model" = lung_model
+  ),
+  #statistic = "mad",
+  fmt = 2
+)
 
@@
 
-pp_check(peak_poisson_model) +
+pp_check(heart_model) +
   theme(legend.position = "bottom")
 
-pp_stan_neg_binom_model <- pp_check(peak_poisson_model)
-pp_stan_neg_binom_model + theme(legend.position = "bottom")
-
-pp_check(peak_neg_binom_model) +
+pp_check(lung_model) +
   theme(legend.position = "bottom")
-
-pp_stan_neg_binom_model <- pp_check(peak_neg_binom_model)
-pp_stan_neg_binom_model + theme(legend.position = "bottom")
 
@@
 plot(first_model, "rhat")
+
+plot(stan_neg_binom_model, "trace")
+
+plot(stan_neg_binom_model, "rhat")
 ```

### 2024-03-14 scripts/00-simulate_data.R

@@ -1,6 +1,6 @@
 #### Preamble ####
 # Purpose: Cleans the two data sets for the years 2018 to 2021
-# Author: Navya Gupta, Shivank Goel, Vanshika Vanshika
+# Author: Navya Hooda, Shivank Goel, Vanshika Vanshika
 # Date: 07 March 2024 
 # Contact: shivankg.goel@mail.utoronto.ca
 # License: MIT

### 2024-03-14 scripts/01-data_cleaning.R

@@ -1,6 +1,6 @@
 #### Preamble ####
 # Purpose: Cleans the two data sets for the years 2018 to 2021
-# Author: Navya Gupta, Shivank Goel, Vanshika Vanshika
+# Author: Navya Hooda, Shivank Goel, Vanshika Vanshika
 # Date: 07 March 2024 
 # Contact: shivankg.goel@mail.utoronto.ca
 # License: MIT
@@ -59,7 +59,7 @@ write.csv(cleaned_data, file = "data/analysis_data/cleaned_air_data.csv", row.na
 
 
 # Filter the data to include only rows where the Cause is "Acute myocardial infarction" or "Malignant neoplasms of trachea, bronchus and lung"
-filtered_df <- df[df$Cause %in% c("Malignant neoplasms of trachea, bronchus and lung"), ]
+filtered_df <- df[df$Cause %in% c("Malignant neoplasms of trachea, bronchus and lung", "Acute myocardial infarction", "Malignant neoplasms of trachea, bronchus and lung", "Other chronic obstructive pulmonary disease", "All other forms of chronic ischemic heart disease"), ]
 
 # Select only the necessary columns: Year, Cause, Ranking, and Total Deaths
 cleaned_df <- filtered_df[, c("Calendar Year", "Cause", "Ranking", "Total Deaths")]
@@ -71,10 +71,10 @@ cleaned_df <- cleaned_df %>%
 cleaned_df = clean_names(cleaned_df)
 
 chart <- read.csv("data/raw_data/chart.csv")
-
+chart
 names(chart)[1] <- "year"
 # Select only the necessary columns: Year, Cause, Ranking, and Total Deaths
-chart <- chart [, c("year", "Provincial.Average")]
+chart <- chart [, c("year", "Provincial.Average", "X10th.Percentile","X90th.Percentile","CAAQS")]
 chart <- chart %>%
   filter(`year` >= 2001)
 chart = clean_names(chart)

### 2024-03-14 scripts/03-model.R

@@ -1,0 +1,77 @@
+#### Preamble ####
+# Purpose: Models different relations between variables such as mortality rates and air pollutants. 
+# Author: Vanshika Vanshika, Shivank Goel, Navya Hooda
+# Date: 13 March 2024
+# Contact: vanshika.vanshika@mail.utoronto.ca
+# License: MIT
+
+#### Workspace setup ####
+library(tidyverse)
+library(rstanarm)
+library(here)
+library(readr)
+library(MASS)
+
+#### Read data ####
+deaths_data <- read.csv(here("data/analysis_data/cleaned_file.csv"))
+air_quality_data <- read.csv(here("data/analysis_data/cleaned_air_data.csv"))
+data_average <- read.csv(here("data/analysis_data/cleaned_chart_data.csv"))
+data_peak <- read.csv(here("data/analysis_data/cleaned_peak_data.csv"))
+alberta_cod <- merge(deaths_data, air_quality_data, by = "year")
+heart <- read.csv(here("data/analysis_data/merged_heart_data.csv"))
+lung <- read.csv(here("data/analysis_data/merged_data.csv"))
+
+# Fit Poisson regression model
+cause_of_death_alberta_poisson <-
+  stan_glm(
+    total_deaths ~ cause,
+    data = alberta_cod,
+    family = poisson(link = "log"),
+    seed = 853
+  )
+
+# Fit Negative Binomial regression model
+cause_of_death_alberta_neg_binomial <-
+  stan_glm(
+    total_deaths ~ cause,
+    data = alberta_cod,
+    family = neg_binomial_2(link = "log"),
+    seed = 853
+  )
+
+
+
+# Convert year and air_quality_health_index to numeric
+heart <- heart %>%
+  mutate(provincial_average = as.numeric(provincial_average),
+         total_deaths = as.numeric(total_deaths))
+lung <- lung %>%
+  mutate(provincial_average = as.numeric(provincial_average),
+         total_deaths = as.numeric(total_deaths))
+
+# Fit models
+
+# Bayesian Negative Binomial Model with stan_glm
+heart_model <- stan_glm(total_deaths ~ provincial_average, data = heart, 
+                                 family = neg_binomial_2(link = "log"), seed = 853)
+
+# Poisson Regression Model
+lung_model <- stan_glm(total_deaths ~ provincial_average, data = lung, 
[…]

### 2024-03-14 scripts/04-model.R

@@ -1,180 +1,0 @@
-# #### Preamble ####
-# # Purpose: Models... [...UPDATE THIS...]
-# # Author: Rohan Alexander [...UPDATE THIS...]
-# # Date: 11 February 2023 [...UPDATE THIS...]
-# # Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
-# # License: MIT
-# # Pre-requisites: [...UPDATE THIS...]
-# # Any other information needed? [...UPDATE THIS...]
-# 
-# 
-# #### Workspace setup ####
-# library(tidyverse)
-# library(rstanarm)
-# 
-# #### Read data ####
-# deaths_data <- read.csv(here("data/analysis_data/cleaned_file.csv"))
-# air_quality_data <- read.csv(here("data/analysis_data/cleaned_air_data.csv"))
-# data_average <- read.csv(here("data/analysis_data/cleaned_chart_data.csv"))
-# data_peak <- read.csv(here("data/analysis_data/cleaned_peak_data.csv"))
-# 
-# ### Model data ####
-# # first_model <-
-# #   stan_glm(
-# #     formula = flying_time ~ length + width,
-# #     data = analysis_data,
-# #     family = gaussian(),
-# #     prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
-# #     prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
-# #     prior_aux = exponential(rate = 1, autoscale = TRUE),
-# #     seed = 853
-# #   )
-# 
-# 
-# library(tidyverse)
-# library(readr)
-# library(MASS)
-# 
-# # Convert year and air_quality_health_index to numeric
-# data <- air_quality_data %>%
-#   mutate(year = as.numeric(year),
-#          air_quality_health_index = as.numeric(air_quality_health_index))
-# 
-# # Fit negative binomial regression model
-# first_model <- glm.nb(air_quality_health_index ~ year, data = data)
-# 
-# 
-# # Fit Poisson regression model
-# poisson_model <- glm(air_quality_health_index ~ year, data = data, family = poisson())
-# 
-# # For the four causes and total deaths
-# # Assuming the dataset for causes and total deaths is named 'causes_data'
-# 
-# causes_model <- glm.nb(total_deaths ~ year + cause, data = deaths_data)
-# 
-# ## model for concentrations
-# 
-# 
-# # # Fit negative binomial regression model using stan_glm
[…]

### 2024-03-14 scripts/99-replications.R

@@ -13,7 +13,7 @@ library(tidyverse)
 # [...UPDATE THIS...]
 
 #### Load data ####
-# [...ADD CODE HERE...]
+# [...ADD CODE HERE...] 
 
 
 

### 2024-03-15 paper/paper.qmd
more changes to the paper

@@
 ---
-title: "Respiration Related Death Cause show Signficant Relation to Increasing "
+title: "Respiratory Related Deaths Show Positive Relation to Increasing Air Pollution"
 subtitle: "My subtitle if needed"
@@
 ```{r}
-#| label: fig-bills
-#| fig-cap: Bills of penguins
+#| label: fig-all
+#| fig-cap: Provincial Average PM2.5 Quantity Vs Total Deaths
 #| echo: false
+#| warning: false
 
@@
   geom_smooth(method = "lm", se = FALSE) +  # Add linear regression line without confidence interval
-  labs(x = "Provincial Average", y = "Total Deaths", title = "Regression Graph") +
+  labs(x = "Provincial Average PM2.5", y = "Total Deaths") +
   theme_minimal()
@@
 ```{r}
-#| label: fig-bills
-#| fig-cap: Bills of penguins
+#| label: fig-lung
+#| fig-cap: Lung Related Causes Mortality Rates vs Average PM2.5 Quantity 
 #| echo: false
+#| warning: false
 
@@
   geom_smooth(method = "lm", se = FALSE) +  # Add linear regression line without confidence interval
-  labs(x = "Provincial Average", y = "Total Deaths", title = "Regression Lung") +
+  labs(x = "Provincial Average", y = "Total Deaths") +
   theme_minimal()
@@
 ```{r}
-#| label: fig-bills
-#| fig-cap: Bills of penguins
+#| label: fig-heart
+#| fig-cap: Heart Related Causes Mortality Rates vs Average PM2.5 Quantity 
 #| echo: false
+#| warning: false
 # Plot the regression graph for all heart causes
@@
   geom_smooth(method = "lm", se = FALSE) +  # Add linear regression line without confidence interval
-  labs(x = "Provincial Average", y = "Total Deaths", title = "Regression Heart") +
+  labs(x = "Provincial Average", y = "Total Deaths") +
   theme_minimal()
@@
 ```{r}
-#| label: fig-planes
-#| fig-cap: Relationship between wing length and width
+#| label: fig-annual
+#| fig-cap: Annual Trends of PM2.5 Concentrations in Alberta
 #| echo: false
@@
   geom_line(aes(y = x90th_percentile, colour = "90th Percentile")) +
-  labs(title = "Annual Trends of PM2.5 Concentrations in Alberta: Provincial Averages and Percentile Ranges (2012-2021)",
-       x = "Years", y = "PM2.5 Concentration (μg/m³)") +
+  labs(x = "Years", y = "PM2.5 Concentration (μg/m³)") +
   scale_colour_manual(values = c("Provincial Average" = "blue", "10th Percentile" = "green", "90th Percentile" = "red")) +
@@
 p
+```
+
+```{r}
+#| label: fig-four
+#| fig-cap: Mortality Rates vs. Air Quality Health Index
+#| echo: false
+#| warning: false
+#| message: false
 
@@
   facet_wrap(~cause) +
-  labs(title = "Scatter Plot of Mortality Rates vs. Air Quality Health Index",
+  labs(
        x = "Average Air Quality Health Index",
        y = "Total Deaths")
+```
 
+```{r}
+#| label: fig-fakemodel
+#| fig-cap: Model data
+#| echo: false
+#| warning: false
+#| message: false
 # Fit the model
@@
 plot(model)
+```
+```{r}
+#| label: fig-fakeprediction
+#| fig-cap: Prediction of deaths based on the model
+#| echo: false
+#| warning: false
+#| message: false
 
@@
   theme_minimal()
+```
 
+```{r}
+#| label: fig-plane
+#| fig-cap: Relationship between wing length and width
+#| echo: false
+#| warning: false
+#| message: false
 # Effect of AQHI on Predicted Deaths
@@
   facet_wrap(~cause)
+```
 
-# Merge datasets on 'year'
-combined_data <- merge(deaths_data, average_aqi_by_year, by = "year")
-
-# Plot scatter plot of mortality rates vs. air quality health index
-ggplot(combined_data, aes(x = average_aqi, y = total_deaths)) +
-  geom_point() +
-  geom_smooth(method = "lm") +
-  facet_wrap(~cause) +
-  labs(title = "Scatter Plot of Mortality Rates vs. Air Quality Health Index",
-       x = "Average Air Quality Health Index",
-       y = "Total Deaths")
+```{r}
+#| label: fig-cause
+#| fig-cap: Total Deaths by Year for Each Cause
+#| echo: false
+#| warning: false
+#| message: false
 
@@
   geom_line() +
-  labs(title = "Total Deaths by Year for Each Disease",
+  labs(
        x = "Year",
-       y = "Total Deaths")
+       y = "Total Deaths") +
+  theme(legend.position = "top") +
+  theme(legend.text = element_text(size = 5)) +
+  guides(color = guide_legend(direction = "vertical"))
+```
 
+```{r}
+#| label: fig-average_air
+#| fig-cap: Average Air Quality Health Index by Year
+#| echo: false
+#| warning: false
+#| message: false
 # Ensure the 'year' column is in numeric format for proper line plotting
@@
   geom_point() + # for points
-  labs(title = "Average Air Quality Health Index by Year",
+  labs(
        x = "Year",
@@
   scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) # Optional: Adjust x-axis breaks
-
+```
+```{r}
+#| label: fig-average_extra
+#| fig-cap: Average Air Quality Health Index by Year
+#| echo: false
+#| warning: false
+#| message: false
 ggplot(average_aqi_by_year, aes(x = as.factor(year), y = average_aqi, fill = as.factor(year))) +
   geom_bar(stat = "identity") +
-  labs(title = "Average Air Quality Health Index by Year",
+  labs(
        x = "Year",
@@
 \begin{align}
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \alpha + \beta_i + \gamma_i\\
-\alpha &\sim \mbox{Normal}(0, 2.5) \\
-\beta &\sim \mbox{Normal}(0, 2.5) \\
-\gamma &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1)
+y_i | \mu_i, \phi &\sim \text{NegBin}(\mu_i, \phi) \\
+\mu_i &= \exp(\alpha + \beta x_i) \\
+\alpha &\sim \text{Normal}(0, 2.5) \\
+\beta &\sim \text{Normal}(0, 2.5) \\
+\phi &\sim \text{Exponential}(1)
 \end{align}
@@
 #| eval: true
-#| label: tbl-modelresults
-#| tbl-cap: "Model of "
+#| label: tbl-modelcauses
+#| tbl-cap: "Different causes of mortality and their death counts"
 #| warning: false
 library(modelsummary)
-
-modelsummary::modelsummary(
-  list(
-    "heart causes model" = heart_model
-  ),
-  #statistic = "mad",
-  fmt = 2
-)
-
-modelsummary::modelsummary(
-  list(
-    "Lung Causes model" = lung_model
-  ),
-  #statistic = "mad",
-  fmt = 2
-)
 
@@
   }
+)
+
+```
+
+```{r}
+#| echo: false
+#| eval: true
+#| label: tbl-modelheart
+#| tbl-cap: "Heart related causes death count and the air quality"
+#| warning: false
+
+modelsummary::modelsummary(
+  list(
+    "heart causes model" = heart_model
+  ),
+  #statistic = "mad",
+  fmt = 2
+)
+
+```
+
+```{r}
+#| echo: false
+#| eval: true
+#| label: tbl-modellung
+#| tbl-cap: "Lung related causes death count and the air quality"
+#| warning: false
+modelsummary::modelsummary(
+  list(
+    "Lung Causes model" = lung_model
+  ),
+  #statistic = "mad",
+  fmt = 2
 )
@@
 #| warning: false
-#| label: fig-ppcheckandposteriorvsprior
+#| label: fig-ppcheckfirst
 #| layout-ncol: 2
@@
   theme(legend.position = "bottom")
+```
+
+```{r}
+#| eval: true
+#| echo: false
+#| message: false
+#| warning: false
+#| label: fig-ppcheckneg
+#| layout-ncol: 2
+#| fig-cap: "Examining how the model fits, and is affected by, the data"
+#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
 
@@
 pp_stan_poisson_model + theme(legend.position = "bottom")
+```
+
+```{r}
+#| eval: true
+#| echo: false
+#| message: false
+#| warning: false
+#| label: fig-ppcheckheartlung
+#| layout-ncol: 2
+#| fig-cap: "Examining how the model fits, and is affected by, the data"
+#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
 

### 2024-03-15 paper/paper.qmd
Introduction of the paper done

@@
 ---
-title: "Respiratory Related Deaths Show Positive Relation to Increasing Air Pollution"
-subtitle: "My subtitle if needed"
+title: "Respiratory-Related Mortality Rates Show A Positive Correlation With Increasing Air pollution"
+subtitle: "Based on Data Collected From The Province Of Alberta"
 author: 
@@
 library(tidyverse)
-library(palmerpenguins)
 library(readxl)
@@
 
-You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.
+The impact of air pollution on human health has increasingly become a global concern, with respiratory illnesses being a significant consequence of poor air quality.
+The National Library of Medicine reports that approximately 4 million people die prematurely each year from chronic respiratory diseases linked to air pollution @ncbi.
+The World Health Organization also suggests that air pollution poses a risk for all-cause mortality and specific diseases, especially the exposure to air pollution is strongly linked with outcomes such as stroke, ischemic heart disease, chronic obstructive pulmonary disease, lung cancer, pneumonia, and cataracts @who.
 
-The remainder of this paper is structured as follows. @sec-data....
+The pollutants most significantly concerning for public health, are particulate matter (PM), carbon monoxide (CO), ozone (O3), nitrogen dioxide (NO2), and sulfur dioxide (SO2).
+Of these, fine particulate matter is particularly worrisome due to its ability to penetrate the lungs deeply, enter the bloodstream, and reach organs, leading to systemic damage to tissues and cells.
+
+Specifically, PM2.5 is of greater concern due to its harmful properties and dangers.
+PM2.5 is airborne particulate matter (PM2.5) and is not a single pollutant but rather a mixture of many chemical species and aerosols.
+PM2.5 is associated with the greatest proportion of adverse health effects related to air pollution, both in the United States and worldwide.
+Long-term (months to years) exposure to PM2.5 has been linked to premature death, particularly in people who have chronic heart or lung diseases, and reduced lung function growth in children @pm.
+
+In this paper, we analyze data from Alberta, Canada, and aim to explore the estimand of the number of deaths that can be correlated to the Air Quality Health Index (AQHI) and PM2.5 quantities.
+Specifically, we assess how AQHI relates to the prevalence of respiratory and cardiac illnesses in this specific region, focusing on four major types: Chronic Obstructive Pulmonary Disease, Ischemic Heart Disease, Acute Myocardial Infarction, and Lung Cancer.
+We use the mortality rate in Alberta data to select respiratory-related illnesses and air quality health index data for Alberta through their provincial open data portal.
+To analyze respiratory illnesses, we use air pollution measured through the AQHI and specifically the particulate matter 2.5 (PM2.5) air pollutant as a predictor of mortality related to respiratory illnesses.
+In total, we will be assessing the overall relationship between AQHI and respiratory illnesses, the relationship between PM2.5 and respiratory illnesses, as well as the relevance of PM2.5 levels in lung and heart disease values as a means to draw any notable significance of PM2.5 in different types of illnesses.
+
+Using negative binomial regression, this study seeks to uncover trends and correlations between AQHI and the prevalence of respiratory illnesses.
+Negative binomial regression is a type of generalized linear model (GLM) that is specifically designed to model count data.
+In our paper, we are using it to predict the number of deaths, based on the year and illness, for the pollutants present in the air.
+By analyzing this relationship, we provide valuable insights that can inform policymakers, healthcare professionals, and the public about the impact of air pollution on respiratory health in Alberta.
+This research aims to contribute to a better understanding of the health effects of air pollution and to support the development of targeted strategies for air quality improvement and public health protection in Alberta.
+Our analytical framework explores whether the variables assessed seem to correlate, if at all, to the total deaths in respiratory-related illnesses through 2012-2022.
+
+This paper is organized as follows: In the Data section, we outline the sources of three different datasets, detail the data-cleaning processes applied to each dataset, and describe any data merging procedures used to prepare the data for input into various models.
+The Results section focuses on analyzing trends and correlations between AQHI, PM2.5, and respiratory and cardiac illnesses.
+Additionally, we discuss the trends and patterns identified by our model, along with the correlation analysis between these variables.
+In the Discussion section, we present our overall findings, discuss any biases and weaknesses in the data that may have influenced these findings, and explain our approach to analyzing these limitations.
+
+Overall, this research has the potential to inform public health strategies and interventions aimed at reducing respiratory illnesses and improving overall health in Alberta and beyond.
 
@@
 
-Some of our data is of penguins (@fig-bills), from @palmerpenguins.
+## Data Source and Collection:
+
+The study relies on datasets obtained from the provincial open databases of Alberta, accessible through the official website @alberta.
+Three key datasets were utilized to extract relevant variables for analysis, aiming to uncover the relationship between air quality and mortality rates in Alberta.
+The analysis begins with the leading causes of death dataset for Alberta, sourced from the provincial open data portal @deaths.
+This dataset provides insights into mortality rates associated with various illnesses, facilitating the examination of trends related to respiratory and heart-related illnesses.
+To explore potential correlations between air quality and mortality rates, the study incorporates the Air Quality Health Index (AQHI) dataset for Alberta, sourced from the provincial open data portal @air.
+This dataset offers comprehensive information on the AQHI across different municipalities in Alberta over multiple years.
+Additionally, the study utilizes PM2.5 air pollutant concentration level data sourced from Alberta's official resources @pollution.
+This dataset provides detailed information on the concentration levels of PM2.5 pollutants over several years, offering valuable insights into air quality trends.
+The following subsections outline the sources, collection methodologies, and data-cleaning procedures implemented to ensure the accuracy and reliability of the datasets used in the analysis.
+This meticulous approach ensures that the data is prepared for thorough analysis, facilitating the exploration of correlations between air quality indicators and mortality rates in Alberta.
+
+Leading Causes of Death in Alberta Data: The disease data is found from the government of Alberta’s open data portal, and was last updated on September 22, 2023 and continues to be updated annually.
+This dataset encompasses mortality data related to the top 30 common causes of death.
+It reports on types of diseases, causes of death, mortality denoted by total death counts, and ranking for 2000-2022.
+Due to our focus on respiratory illnesses, in the leading cause of death dataset, we grouped diseases by categories.
+Our category of focus included filtering on illnesses like acute myocardial infarction, malignant neoplasms of the trachea, bronchus, and lung, other chronic obstructive pulmonary disease, and all other forms of chronic ischemic heart disease.
+Leading causes of death are measured and ranked by the top 30 most common death causes each specific year.
+The causes of death are classified based on the International Classification of Diseases 10th Edition.
+
+AQHI Data: The second dataset we used is the air quality health index (AQHI) dataset found at the government of Alberta’s open data portal.
+This dataset contains AQHI by municipality for the years 2012-2022 and reports air quality health index, and health risk both quantitatively and qualitatively.
+To use the AQHI dataset we employed simple data-cleaning practice to maintain descriptive variable names and readability.
+The data is measured by the percentage of hours for each year at a given air quality level, by municipality.
+The Air Quality Health Index is calculated based on the relative risks of a combination of common air pollutants that is known to harm human health.
+These pollutants are ozone (O3) at ground level, particulate matter (PM2.5), and nitrogen dioxide (NO2).
+Risks are defined as follows: 1-3 High Quality; 4-6 Moderate Quality; 7-9 Low Quality; 10+ Very Low Quality.
+
+PM2.5 Data: We used the PM2.5 data set retrieved from Alberta.ca (Government of Alberta) which was last updated in April 2023.
+It reports on average PM2.5 concentration levels through the years 2000-2021 using a provincial average, the 10th percentile quantities, and the 90th percentile quantities, with a focus on 8 municipalities Edmonton, Fort McMurray, Grande Prairie, Lethbridge, Medicine Hat, and Red Deer respectively and lastly reports the Canadian Ambient Air Quality Standard (CAAQS) value.
+
+The Alberta Air Zone report @report, which is linked to our dataset, provides a detailed explanation of the measurement and processing of the PM2.5 quantity.
+Alberta Air Zones divides Alberta into six air zones which are aligned with Alberta’s Land-use Framework regional boundaries.
+Ambient air quality in Alberta is monitored at continuous air monitoring stations located within these air zones.
+PM2.5 quantities are taken throughout these stations across Alberta, and they measure the quantities in µg (micrograms per cubic meter of air).
+
+## Data Cleaning
+
+We used R [@citeR] and @rohan for data cleaning and processing, utilizing packages like tidyverse [@tidy] for data manipulation and janitor [@jan] for cleaning column names.
+Other packages used includes `ggplot2` [@ggplot], `dplyr` [@dp], `readr` [@read], `tibble` [@tib], `janitor` [@jan],`reshape2` [@reshape], `knitr` [@knit], `ggbeeswarm` [@gg], `ggrepel` [@repel], `kableExtra`[@kable], `readxl`[@readxl], `MASS`[@mass], `rstanarm`[@rstan], `modelsummary`[@model] and `here` [@here].
+
+The raw air quality data were preprocessed to remove inconsistencies and irrelevant information.
+Specifically, we filtered the dataset to include observations from the years 2012 to 2021, which are relevant to our analysis.
+Additionally, we merged this dataset with additional information on peak pollution levels for comprehensive analysis.
+Similar to the previous datasets, the raw mortality data underwent cleaning procedures to focus on specific causes relevant to our analysis.
+We filtered the dataset to include observations up to 2021 and merged it with additional information on air quality for correlation analysis.
+The raw data on AQHI were filtered to include observations from the years 2011 to 2021 for consistency with other datasets.
+Additionally, the data were aggregated at the municipal level for further analysis.
+
+## Data Modifications
+
+In this study, we constructed unique datasets by thoughtfully selecting and merging data from the Government of Alberta's open data portal, Alberta.ca, spanning the years 2012 to 2022.
+Our process involved merging variables from various datasets to create specific datasets tailored for model building and analysis.
+One such dataset, 'cleaned_chart_data,' was created by merging variables such as causes of death, total deaths, provincial average PM2.5 levels, and CAAQS.
+This dataset was designed to facilitate our analysis of any significant correlations between these variables.
+A snapshot of this data is referenced in Table X.
+Additionally, we derived two other datasets, 'merged_data' and 'merged_heart_data,' by merging variables related to heart disease numbers, lung disease numbers, and provincial average PM2.5 values.
+These datasets were instrumental in examining the impact of PM2.5 on each type of illness, as previously discussed.
+Overall, our methodology ensured the creation of comprehensive datasets that allowed for a detailed investigation into the relationships between PM2.5 levels and various health outcomes in Alberta.
 
@@
 ```{r}
-#| label: fig-fakemodel
-#| fig-cap: Model data
-#| echo: false
-#| warning: false
-#| message: false
-# Fit the model
-model <- glm.nb(total_deaths ~ average_aqi + factor(year) + factor(cause), data = combined_data)
-
-# Model Diagnostics
-par(mfrow=c(2,2))
-plot(model)
-```
-```{r}
-#| label: fig-fakeprediction
-#| fig-cap: Prediction of deaths based on the model
-#| echo: false
-#| warning: false
-#| message: false
-
-# Add Predictions to the combined data
-# Note: predict() function might throw an error if there are new factor levels in the test set that were not present in the training set
-# To prevent this, the newdata argument should only include levels that were present in the model fitting
-combined_data$predicted_deaths <- predict(model, newdata = combined_data, type = "response")
-
-# Plot Observed vs. Predicted Deaths
-ggplot(combined_data, aes(x = total_deaths, y = predicted_deaths)) +
-  geom_point() +
-  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
-  labs(x = "Observed Deaths", y = "Predicted Deaths") +
-  theme_minimal()
-```
-
-```{r}
-#| label: fig-plane
-#| fig-cap: Relationship between wing length and width
-#| echo: false
-#| warning: false
-#| message: false
-# Effect of AQHI on Predicted Deaths
-ggplot(combined_data, aes(x = average_aqi, y = predicted_deaths, color = factor(year))) +
-  geom_point() +
-  geom_line(aes(group = factor(year))) + # Adding lines to connect points by year
-  labs(x = "Average Air Quality Health Index", y = "Predicted Deaths") +
-  theme_minimal() +
-  facet_wrap(~cause)
-```
-
-```{r}
 #| label: fig-cause
@@
 ```
-```{r}
-#| label: fig-average_extra
-#| fig-cap: Average Air Quality Health Index by Year
-#| echo: false
-#| warning: false
-#| message: false
-ggplot(average_aqi_by_year, aes(x = as.factor(year), y = average_aqi, fill = as.factor(year))) +
-  geom_bar(stat = "identity") +
-  labs(
-       x = "Year",
-       y = "Average AQHI") +
-  theme_minimal()
-
-```
-
 Talk way more about it.
@@
 
+# # Make predictions for heart and lung models
+# heart_predictions <- predict(heart_model, newdata = data.frame(provincial_average = heart$provincial_average), se.fit = TRUE)
+# lung_predictions <- predict(lung_model, newdata = data.frame(provincial_average = lung$provincial_average), se.fit = TRUE)
+# # Create trend graphs
+# # For example, let's plot predicted values against provincial average for heart data
+# heart_pred_df <- data.frame(provincial_average = heart$provincial_average,
+#                             total_deaths_pred = heart_predictions$fit,
+#                             total_deaths_pred_upper = heart_predictions$fit + 1.96 * heart_predictions$se.fit,
+#                             total_deaths_pred_lower = heart_predictions$fit - 1.96 * heart_predictions$se.fit)
+# 
+# # Plot
+# ggplot(heart_pred_df, aes(x = provincial_average, y = total_deaths_pred)) +
+#   geom_line() +
+#   geom_ribbon(aes(ymin = total_deaths_pred_lower, ymax = total_deaths_pred_upper), alpha = 0.2) +
+#   labs(x = "Provincial Average", y = "Predicted Total Deaths", title = "Predicted Total Deaths vs. Provincial Average (Heart Data)") +
+#   theme_minimal()
+
+# class(heart_model$provincial_average)
+# class(lung_model$provincial_average)
+# 
+# # Create a range of values for the predictor variable
+# provincial_average_range <- seq(min(heart$provincial_average, lung$provincial_average),
+#                                 max(heart$provincial_average, lung$provincial_average),
+#                                 length.out = 100)
+# 
+# # Make predictions for heart and lung models
+# heart_predictions <- predict(heart_model, newdata = data.frame(provincial_average = provincial_average_range), se.fit = TRUE)
+# lung_predictions <- predict(lung_model, newdata = data.frame(provincial_average = provincial_average_range), se.fit = TRUE)
+# 
+# # Create data frames for plotting
+# heart_pred_df <- data.frame(provincial_average = provincial_average_range,
+#                             total_deaths_pred = heart_predictions$fit,
+#                             total_deaths_pred_upper = heart_predictions$fit + 1.96 * heart_predictions$se.fit,
+#                             total_deaths_pred_lower = heart_predictions$fit - 1.96 * heart_predictions$se.fit)
+# 
+# lung_pred_df <- data.frame(provincial_average = provincial_average_range,
+#                            total_deaths_pred = lung_predictions$fit,
+#                            total_deaths_pred_upper = lung_predictions$fit + 1.96 * lung_predictions$se.fit,
+#                            total_deaths_pred_lower = lung_predictions$fit - 1.96 * lung_predictions$se.fit)
+# 
+# # Plot predicted deaths for heart and lung models
+# ggplot() +
+#   geom_line(data = heart_pred_df, aes(x = provincial_average, y = total_deaths_pred, color = "Heart Data")) +
+#   geom_ribbon(data = heart_pred_df, aes(x = provincial_average, ymin = total_deaths_pred_lower, ymax = total_deaths_pred_upper, fill = "Heart Data"), alpha = 0.2) +
+#   geom_line(data = lung_pred_df, aes(x = provincial_average, y = total_deaths_pred, color = "Lung Data")) +
+#   geom_ribbon(data = lung_pred_df, aes(x = provincial_average, ymin = total_deaths_pred_lower, ymax = total_deaths_pred_upper, fill = "Lung Data"), alpha = 0.2) +
+#   labs(x = "Provincial Average", y = "Predicted Total Deaths", title = "Predicted Total Deaths due to Air Pollution") +
+#   scale_color_manual(values = c("Heart Data" = "blue", "Lung Data" = "red")) +
+#   scale_fill_manual(values = c("Heart Data" = "blue", "Lung Data" = "red"), guide = FALSE) +
+#   theme_minimal()
+
 ```
@@
     "causeMalignant neoplasms of trachea, bronchus and lung" = "Trachea/Bronchus/Lung Cancer",
-    "causeOrganic dementia" = "Organic Dementia",
+    "Acute myocardial infarction" = "Heart Attack",
     "causeOther chronic obstructive pulmonary disease" = "COPD"
@@
 #| warning: false
-#| label: fig-ppcheckneg
-#| layout-ncol: 2
-#| fig-cap: "Examining how the model fits, and is affected by, the data"
-#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
-
-pp_stan_neg_binom_model <- pp_check(stan_neg_binom_model)
-pp_stan_neg_binom_model + theme(legend.position = "bottom")
-
-pp_stan_poisson_model <- pp_check(stan_poisson_model)
-pp_stan_poisson_model + theme(legend.position = "bottom")
-```
-
-```{r}
-#| eval: true
-#| echo: false
-#| message: false
-#| warning: false
 #| label: fig-ppcheckheartlung

### 2024-03-15 scripts/03-model.R

@@ -59,7 +59,15 @@ heart_model <- stan_glm(total_deaths ~ provincial_average, data = heart,
 lung_model <- stan_glm(total_deaths ~ provincial_average, data = lung, 
                        family = neg_binomial_2(link = "log"), seed = 853)
 
+# Convert 'cause' variable to factor with levels
+alberta_cod$cause <- factor(alberta_cod$cause)
 
+# Check the levels of the 'cause' variable
+levels(alberta_cod$cause)
+
+
+# Check the structure of the alberta_cod dataset
+str(alberta_cod)
 #### Save model ####
 saveRDS(neg_binom_model, "models/neg_binom_model.rds")
 saveRDS(stan_neg_binom_model, "models/stan_neg_binom_model.rds")

### 2024-03-16 scripts/02-data_cleaning.R
Delete scripts/02-data_cleaning.R

@@ -1,54 +1,0 @@
-#### Preamble ####
-# Purpose: Cleans the two data sets for the years 2018 to 2021
-# Author: Navya Hooda, Shivank Goel, Vanshika Vanshika
-# Date: 07 March 2024 
-# Contact: shivankg.goel@mail.utoronto.ca
-# License: MIT
-
-#### Workspace setup ####
-library(tidyverse)
-library(janitor)
-library(readxl)
-library(dplyr)
-
-#### Clean data ####
-# first dataset
-df <- read_csv("data/raw_data/deaths-leading-causes.csv", skip = 2)
-
-# Filter the data to include only rows where the Cause is "Acute myocardial infarction" or "Malignant neoplasms of trachea, bronchus and lung"
-filtered_df <- df[df$Cause %in% c("Acute myocardial infarction", "Malignant neoplasms of trachea, bronchus and lung", "Other chronic obstructive pulmonary disease", "All other forms of chronic ischemic heart disease"), ]
-
-# Select only the necessary columns: Year, Cause, Ranking, and Total Deaths
-cleaned_df <- filtered_df[, c("Calendar Year", "Cause", "Ranking", "Total Deaths")]
-
-# Rename the columns
-colnames(cleaned_df) <- c("Year", "Cause", "Ranking", "Total Deaths")
-
-cleaned_df <- cleaned_df %>%
-  filter(`Year` >= 2012)
-cleaned_df = clean_names(cleaned_df)
-
-# Write the cleaned data to a new CSV file
-write_csv(cleaned_df, "data/analysis_data/cleaned_file.csv")
-
-
-## Other dataset ##
-
-data <- read_csv("data/raw_data/download.csv")
-
-# Select only the necessary columns and rename them
-cleaned_data <- data[, c("CSD", "Period", "Air Quality Health Index", "Health Risk", "OriginalValue")]
-# Rename the columns
-cleaned_data <- cleaned_data %>%
-  rename(Municipality = CSD,
-         Year = Period)
-
-cleaned_data <- cleaned_data %>%
-  arrange(Year)
-cleaned_data <- cleaned_data %>%
-  filter(`Year` >= 2012)
-
-cleaned_data = clean_names(cleaned_data)
-
-# Save the cleaned dataset to a CSV file
-write.csv(cleaned_data, file = "data/analysis_data/cleaned_air_data.csv", row.names = FALSE)

### 2024-03-16 paper/paper.qmd
Datasheet and model in the paper

@@
 The study relies on datasets obtained from the provincial open databases of Alberta, accessible through the official website @alberta.
-Three key datasets were utilized to extract relevant variables for analysis, aiming to uncover the relationship between air quality and mortality rates in Alberta.
+Three essential datasets were employed to extract relevant variables for analysis, with the goal of revealing the association between air quality and mortality rates in Alberta.
 The analysis begins with the leading causes of death dataset for Alberta, sourced from the provincial open data portal @deaths.
@@
 To explore potential correlations between air quality and mortality rates, the study incorporates the Air Quality Health Index (AQHI) dataset for Alberta, sourced from the provincial open data portal @air.
-This dataset offers comprehensive information on the AQHI across different municipalities in Alberta over multiple years.
+This dataset offers entensive information on the AQHI across different municipalities in Alberta over multiple years.
 Additionally, the study utilizes PM2.5 air pollutant concentration level data sourced from Alberta's official resources @pollution.
@@
 
-Leading Causes of Death in Alberta Data: The disease data is found from the government of Alberta’s open data portal, and was last updated on September 22, 2023 and continues to be updated annually.
+**Leading Causes of Death in Alberta Data**: The disease data is found from the government of Alberta’s open data portal @alberta, and was last updated on September 22, 2023 and continues to be updated annually.
 This dataset encompasses mortality data related to the top 30 common causes of death.
@@
 
-AQHI Data: The second dataset we used is the air quality health index (AQHI) dataset found at the government of Alberta’s open data portal.
+**AQHI Data**: The second dataset we used is the air quality health index (AQHI) dataset found at the government of Alberta’s open data portal @alberta.
 This dataset contains AQHI by municipality for the years 2012-2022 and reports air quality health index, and health risk both quantitatively and qualitatively.
@@
 
-PM2.5 Data: We used the PM2.5 data set retrieved from Alberta.ca (Government of Alberta) which was last updated in April 2023.
+**PM2.5 Data**: We used the PM2.5 data set retrieved from @alberta which was last updated in April 2023.
 It reports on average PM2.5 concentration levels through the years 2000-2021 using a provincial average, the 10th percentile quantities, and the 90th percentile quantities, with a focus on 8 municipalities Edmonton, Fort McMurray, Grande Prairie, Lethbridge, Medicine Hat, and Red Deer respectively and lastly reports the Canadian Ambient Air Quality Standard (CAAQS) value.
@@
 
-We used R [@citeR] and @rohan for data cleaning and processing, utilizing packages like tidyverse [@tidy] for data manipulation and janitor [@jan] for cleaning column names.
+We used R [@citeR] for data cleaning and processing, utilizing packages like tidyverse [@tidy] for data manipulation and janitor [@jan] for cleaning column names.
 Other packages used includes `ggplot2` [@ggplot], `dplyr` [@dp], `readr` [@read], `tibble` [@tib], `janitor` [@jan],`reshape2` [@reshape], `knitr` [@knit], `ggbeeswarm` [@gg], `ggrepel` [@repel], `kableExtra`[@kable], `readxl`[@readxl], `MASS`[@mass], `rstanarm`[@rstan], `modelsummary`[@model] and `here` [@here].
@@
 Specifically, we filtered the dataset to include observations from the years 2012 to 2021, which are relevant to our analysis.
-Additionally, we merged this dataset with additional information on peak pollution levels for comprehensive analysis.
+Additionally, we merged this dataset with additional information on peak pollution levels for analysis.
+
 Similar to the previous datasets, the raw mortality data underwent cleaning procedures to focus on specific causes relevant to our analysis.
 We filtered the dataset to include observations up to 2021 and merged it with additional information on air quality for correlation analysis.
+
 The raw data on AQHI were filtered to include observations from the years 2011 to 2021 for consistency with other datasets.
@@
 These datasets were instrumental in examining the impact of PM2.5 on each type of illness, as previously discussed.
-Overall, our methodology ensured the creation of comprehensive datasets that allowed for a detailed investigation into the relationships between PM2.5 levels and various health outcomes in Alberta.
-
-```{r}
-#| label: fig-all
-#| fig-cap: Provincial Average PM2.5 Quantity Vs Total Deaths
-#| echo: false
-#| warning: false
-
-# Plot the regression graph
-ggplot(data, aes(x = provincial_average, y = total_deaths)) +
-  geom_point() +
-  geom_smooth(method = "lm", se = FALSE) +  # Add linear regression line without confidence interval
-  labs(x = "Provincial Average PM2.5", y = "Total Deaths") +
-  theme_minimal()
-
-```
+Overall, our methodology ensured the creation of merged datasets that allowed for a detailed investigation into the relationships between PM2.5 levels and various health outcomes in Alberta.
 
@@
   guides(color = guide_legend(direction = "vertical"))
+
 ```
@@
 ```
-Talk way more about it.
 
@@
 
-The goal of our modelling strategy is twofold. Firstly,...
+## Model Set-up
 
-Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
+In our analysis, we implemented a series of regression models to study the impact of air pollution on various health outcomes, especially focusing on respiratory and cardiac illnesses in Alberta, Canada.
+The models used include three variations of negative binomial regression and one Poisson regression model.
 
-## Model set-up
+We used *Negative Binomial Regression (NBR)* to study how air pollution affects deaths related to breathing problems.
+NBR is extremely useful for handling overdispersed count data, where the variance exceeds the mean @neg.
+This method is good for analyzing data where the numbers (like the total number of deaths) vary more than usual.
+In many real-world scenarios, these numbers can be more unpredictable than what simpler methods like the Poisson distribution can handle.
+Negative Binomial Regression works better in these kinds of situations since it accounts for this extra unpredictability.
+It gives us easy-to-understand results that show how average air pollution in a province *(provincial_average)* can affect the number of deaths *(total_deaths)*.
 
-Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.
+Our first model compares different causes of mortality and their corresponding death count.
+The second model investigates the relationship between heart-related causes of death and PM2.5 levels.
+And lastly, the third model analyzes the connection between lung-related causes of death and air quality.
+To build our model for heart and lung-related causes of death, we identified four key diseases from the dataset and grouped them into two broader categories: heart diseases and lung diseases.
+
+The diseases and their grouping are as follows:
+
+*Heart Diseases*:
+
+Ischemic Heart Disease (All other forms of chronic ischemic heart disease)
+
+Heart Attack (Acute Myocardial Infraction)
+
+*Lung Diseases*:
+
+Trachea/Bronchus/Lung Cancer (Malignant neoplasms of trachea, bronchus, and lung)
+
+COPD (Other chronic obstructive pulmonary disease)
+
+In our approach, the total deaths from Ischemic Heart Disease and Heart Attack were summed up to create the dataset for heart diseases, and similarly, data from Trachea/Bronchus/Lung Cancer and COPD were combined for lung diseases.
+
+*Details of Regression Models*
+
+The general form of our negative binomial model is represented as follows.
 
@@
 
-We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.
+1. *Modelling the count data*:
 
-### Model justification
+$y_i|\mu_i, \phi \sim NegBin(\mu_i, \phi)$
 
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
+Here, $y_i$ represents the count of deaths for each health outcome (like heart or lung disease). The model assumes that the count data follow a Negative Binomial distribution.
 
-We can use maths by including latex between dollar signs, for instance $\theta$.
+2. *Link function and predictors*
 
-# Results
+$\mu_i = exp(\alpha + \beta x_i)$
 
-Our results are summarized in @tbl-modelresults.
+Link Function – This is the link function used for the negative binomial regression.
+By default, when we specify dist = negbin, the log link function is assumed (and does not need to be specified) @neg.
+In this equation $\mu_i$  is the expected count of deaths, which we model as an exponential function of the predictors.
+This ensures that our predictions for the count data are always positive.
+The predictor ($x_i$) in our models include average air pollution levels (provincial_average).
 
+3. *Coefficient*
+
+Coefficients are the estimated negative binomial regression coefficients for the model.
+We can interpret the negative binomial regression coefficient as follows: for a one unit change in the predictor variable, the log of expected counts of the response variable changes by the respective regression coefficient, given the other predictor variables in the model are held constant @neg.
+
+$\alpha \sim$ Normal(0,2.5): The intercept $\alpha$ has a normal prior, indicating about our initial assumption about the baseline level of health outcomes in the absence of predictors.
+
+$\beta \sim$ Normal(0,2.5): The coefficients $\beta$  for our predictors also follow a normal distribution.
+These coefficients tells us about the relationship strength between air pollution and the health outcomes.
+
+4. *Dispersion parameter*
+
+As the dispersion parameter gets larger and larger, the variance converges to the same value as the mean, and the negative binomial converges to a Poisson distribution @clay
+$\phi$ is modeled using an exponential distribution. It accounts for extra variation in the count data that is not explained by the Poisson model alone.
+
+## Model Summary and Model Results
+
+*Model 1: Comparative Analysis of Mortality Causes*
+
+Interpretation  of table results from [@tbl-modelcauses]
+
+[@tbl-modelcauses] presents the results from our first model, and compares various causes of mortality using both Poisson and negative binomial regression approaches.
+This table helps in understanding the relationship between air pollution and mortality due to different causes.
+
+1.
+Coefficients - As described above, the coefficients from these models (the $\beta$ values) tell us how changes in the predictors (like provincial average PM2.5 levels) are associated with changes in the response variable (total death counts).
+A positive coefficient indicates that an increase in the predictor leads to an increase in the response, while a negative coefficient suggests the opposite.
+
+*Ischemic Heart Disease*: 
+Both Poisson and negative binomial models showed a coefficient of 0.510, suggesting a consistent and significant association between air pollution and mortality due to Ischemic Heart Disease.
+This positive coefficient highlights a serious public health concern, suggesting that as air pollution worsens, the risk of death due to Ischemic Heart Disease increases.
+
+*Trachea/Bronchus/Lung Cancer*: Similarly, for Trachea/Bronchus/Lung Cancer, both models indicated a coefficient of 0.353, describing the impact of air pollution on these lung diseases.
+It is learned through this coefficient that increased levels of air pollution are associated with higher mortality rates from these types of lung cancer.
+
+*COPD*: For COPD, the coefficient was minimal (0.004), which suggests that, there is a negligible direct association between air pollution and COPD mortality.
+It's important to note that this does not necessarily mean air pollution has no effect on COPD but rather, it might indicate that the relationship is more complex and possibly influenced by other variables not captured in the model.
+
+2. Number of Observations (Num.Obs.):
+
+"Num.Obs. 9048" indicates the total number of data points (observations) used in the model. A higher number of observations often produces better predictions and thus more accuaracy in results.
+
+3.
+Log-Likelihood (Log.Lik.): The log-likelihood values (-69,064.136 for Poisson and -53,402.269 for negative binomial) measure how well the model fits the data.
+A higher (less negative) log-likelihood value generally indicates a better model fit @ibm.
+In this case, the higher log-likelihood for the negative binomial model suggests it fits the data better than the Poisson model, which is consistent with the expectation that the negative binomial model handles overdispersion more effectively.
+This result is also evident using posterior predictive checks [@fig-ppcheckfirst], to show that the negative binomial approach is a better choice for this circumstance.
+
+4.
+Expected Log Predictive Density (ELPD): ELPD (-69,078.6 for Poisson and -53,405.5 for negative binomial) suggests that negative binomial regression may be better at predicting new data compared to the Poisson model, which truly matches our expectation @telling.
+
+5. Leave-One-Out Cross-Validation Information Criterion (LOOIC): The negative binomial model's lower LOOIC suggests it is a better model in terms of prediction and handling the data complexity.
+
+6. Root Mean Square Error (RMSE): RMSE (97.30 for both models) measures the model's prediction error. 
+Root mean square indicates the quality of model fit. Lesser RMSE indicates better model fit and higher RMSE indicates poor model fit @many.
+The fact that RMSE is the same for both models suggests similar predictive accuracy in terms of the magnitude of errors.
 ```{r}
@@
   readRDS(file = here::here("models/poisson_model.rds"))
-
-# # Make predictions for heart and lung models
-# heart_predictions <- predict(heart_model, newdata = data.frame(provincial_average = heart$provincial_average), se.fit = TRUE)
-# lung_predictions <- predict(lung_model, newdata = data.frame(provincial_average = lung$provincial_average), se.fit = TRUE)
-# # Create trend graphs
-# # For example, let's plot predicted values against provincial average for heart data
-# heart_pred_df <- data.frame(provincial_average = heart$provincial_average,
-#                             total_deaths_pred = heart_predictions$fit,
-#                             total_deaths_pred_upper = heart_predictions$fit + 1.96 * heart_predictions$se.fit,
-#                             total_deaths_pred_lower = heart_predictions$fit - 1.96 * heart_predictions$se.fit)
-# 
-# # Plot
-# ggplot(heart_pred_df, aes(x = provincial_average, y = total_deaths_pred)) +
-#   geom_line() +
-#   geom_ribbon(aes(ymin = total_deaths_pred_lower, ymax = total_deaths_pred_upper), alpha = 0.2) +
-#   labs(x = "Provincial Average", y = "Predicted Total Deaths", title = "Predicted Total Deaths vs. Provincial Average (Heart Data)") +
-#   theme_minimal()
-
-# class(heart_model$provincial_average)
-# class(lung_model$provincial_average)
-# 
-# # Create a range of values for the predictor variable
-# provincial_average_range <- seq(min(heart$provincial_average, lung$provincial_average),
-#                                 max(heart$provincial_average, lung$provincial_average),
-#                                 length.out = 100)
-# 
-# # Make predictions for heart and lung models
-# heart_predictions <- predict(heart_model, newdata = data.frame(provincial_average = provincial_average_range), se.fit = TRUE)
-# lung_predictions <- predict(lung_model, newdata = data.frame(provincial_average = provincial_average_range), se.fit = TRUE)
-# 
-# # Create data frames for plotting
-# heart_pred_df <- data.frame(provincial_average = provincial_average_range,
-#                             total_deaths_pred = heart_predictions$fit,
-#                             total_deaths_pred_upper = heart_predictions$fit + 1.96 * heart_predictions$se.fit,
-#                             total_deaths_pred_lower = heart_predictions$fit - 1.96 * heart_predictions$se.fit)
-# 
-# lung_pred_df <- data.frame(provincial_average = provincial_average_range,
-#                            total_deaths_pred = lung_predictions$fit,
-#                            total_deaths_pred_upper = lung_predictions$fit + 1.96 * lung_predictions$se.fit,
-#                            total_deaths_pred_lower = lung_predictions$fit - 1.96 * lung_predictions$se.fit)
-# 
-# # Plot predicted deaths for heart and lung models
-# ggplot() +
-#   geom_line(data = heart_pred_df, aes(x = provincial_average, y = total_deaths_pred, color = "Heart Data")) +
-#   geom_ribbon(data = heart_pred_df, aes(x = provincial_average, ymin = total_deaths_pred_lower, ymax = total_deaths_pred_upper, fill = "Heart Data"), alpha = 0.2) +
-#   geom_line(data = lung_pred_df, aes(x = provincial_average, y = total_deaths_pred, color = "Lung Data")) +
-#   geom_ribbon(data = lung_pred_df, aes(x = provincial_average, ymin = total_deaths_pred_lower, ymax = total_deaths_pred_upper, fill = "Lung Data"), alpha = 0.2) +
-#   labs(x = "Provincial Average", y = "Predicted Total Deaths", title = "Predicted Total Deaths due to Air Pollution") +
-#   scale_color_manual(values = c("Heart Data" = "blue", "Lung Data" = "red")) +
-#   scale_fill_manual(values = c("Heart Data" = "blue", "Lung Data" = "red"), guide = FALSE) +
-#   theme_minimal()
 
@@
 ```
+*Model 2: Heart Disease and Air Quality*
+
+Model 2 discusses and provides results about the specific relationship between heart-related diseases and PM2.5 levels.
+
+Interpretation of table results from [@tbl-modelheart]:
+
+[@tbl-modelheart] showcases the relationship between heart-related diseases and PM2.5 levels, utilizing our second regression model.
+
+1.
+Intercept (8.02): The intercept value of 8.02 represents the log count of heart disease-related deaths when the PM2.5 level is at its baseline (zero).
+This high value suggests other influential factors for heart disease mortality beyond air pollution levels.
+
+2.
+Provincial Average PM2.5 (Coefficient = 0.00): The coefficient for provincial average PM2.5 being 0.00 indicates no significant impact of PM2.5 levels on heart disease mortality within the analyzed range.
+This finding shows that heart disease is caused by many factors, so public health plans should use various methods to address it.
+
+3. Number of Observations: The analysis is based on a smaller dataset of 21 observations to study particular relationship over the years between heart diseases and PM2.5.
+
+4.
+Model Fit and Predictive Accuracy: Key statistical measures (Log-Likelihood, ELPD, LOOIC, RMSE), similar to model 1, would be helpful in edetermining the model's fit and predictive power.
+A higher Log-Likelihood, lower LOOIC, and comparable RMSE with other models would further validate the findings.
 
@@
 
+*Model 3: Lung Disease and Air Quality*
+
+Interpretation of table results from [@tbl-modellung]
+
+[@tbl-modellung] focuses on the connection between lung-related diseases and air quality, as represented by PM2.5 levels.
+
+1.
+Intercept (7.57): The intercept of 7.57, indicates the baseline level of lung disease mortality in the absence of PM2.5 contributions.
+This again suggest the influence of other variables in lung disease outcomes.
+
+2. Provincial Average PM2.5 (Coefficient = 0.01): A small but positive coefficient for PM2.5 suggests a marginal increase in lung disease mortality with rising PM2.5 levels.
+
+3. Number of Observations: As with the previous model 2, the number of observations were 21 to study particular relationship over the years between lung diseases and PM2.5.
+
+4.
+Model Fit and Predictive Accuracy: Evaluating this model using Log-Likelihood, ELPD, LOOIC, and RMSE is crucial.
+For instance, a higher Log-Likelihood and lower LOOIC compared to alternative models would indicate a better fit for the lung disease data.
+
 ```{r}
@@
 ```
+
+# Results
+
+Our results are summarized in @tbl-modelresults.
+
 # Discussion
@@
 
-plot(stan_neg_binom_model, "trace")
+plot(lung_model, "trace")
 
-plot(stan_neg_binom_model, "rhat")
+plot(lung_model, "rhat")
+
+plot(heart_model,"trace" )
+
+plot(heart_model, "rhat")
 ```

### 2024-03-16 scripts/02-test_data.R

@@ -1,15 +1,70 @@
 #### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Data Validation and Testing for Mortality and Air Quality Data Analysis
+# Authors: Navya Hooda, Shivank Goel, Vanshika Vanshika
+# Date: 12th February 2024
+# Contact: shivankg.goel@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
+# Pre-requisites: tidyverse, testthat
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
+library(testthat)
+
+
+#Unit Testing Reference:https://testthat.r-lib.org/reference/test_that.html 
+
+# Load the datasets
+cleaned_air_data <- read.csv("data/analysis_data/cleaned_air_data.csv")
+cleaned_avg_pm25_conc <- read.csv("data/analysis_data/cleaned_chart_data.csv")
+cleaned_mortality_data <- read.csv("data/analysis_data/cleaned_file.csv")
 
 #### Test data ####
+
+# Test the structure and type of the air quality data
+test_that("Air quality data columns are of the correct type", {
+  expect_type(cleaned_air_data$year, "integer")
+  expect_type(cleaned_air_data$health_risk, "character")
+  expect_type(cleaned_air_data$original_value, "double")
+})
+
+
+# Test the structure and type of the PM2.5 concentration data
+test_that("PM2.5 concentration data columns are of the correct type", {
+  expect_type(cleaned_avg_pm25_conc$year, "integer")
+  expect_type(cleaned_avg_pm25_conc$provincial_average, "double")
+  expect_type(cleaned_avg_pm25_conc$x10th_percentile, "double")
+  expect_type(cleaned_avg_pm25_conc$x90th_percentile, "double")
+})
+
+# Test the structure and type of the mortality data
+test_that("Mortality data columns are of the correct type", {
+  expect_type(cleaned_mortality_data$year, "integer")
+  expect_type(cleaned_mortality_data$cause, "character")
+  expect_type(cleaned_mortality_data$ranking, "integer")
+  expect_type(cleaned_mortality_data$total_deaths, "integer")
+})
+
[…]

### 2024-03-16 scripts/99-replications.R

@@ -1,19 +1,0 @@
-#### Preamble ####
-# Purpose: Replicated graphs from... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
-
-#### Workspace setup ####
-library(tidyverse)
-# [...UPDATE THIS...]
-
-#### Load data ####
-# [...ADD CODE HERE...] 
-
-
-

### 2024-03-16 paper/paper.qmd
Final paper

@@
 ---
-title: "Respiratory-Related Mortality Rates Show A Positive Correlation With Increasing Air pollution"
+title: "Respiratory-Related Mortality Rates Show A Positive Correlation With Increasing Air Pollution"
 subtitle: "Based on Data Collected From The Province Of Alberta"

### 2024-03-16 paper/paper.qmd
Final Paper

@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "Air pollution's severe health impact on human respiratory and cardiac systems is a growing public health issue.
+The fine particulate matter (PM2.5) arise a primary concern due to its deep lung penetration and systemic effects.
+This research uses data from Alberta, Canada, to study the connection between PM2.5 levels as measured by the Air Quality Health Index (AQHI),  and the number of deaths from certain heart and lung diseases.
+We utilized statistical models, and identified a strong association between increased PM2.5 levels and higher mortality from these diseases.
+These findings suggest that there is an immediate need for health policies that address air quality improvements to safeguard public health for Alberta and beyond."
 format: pdf
@@
 Due to our focus on respiratory illnesses, in the leading cause of death dataset, we grouped diseases by categories.
-Our category of focus included filtering on illnesses like acute myocardial infarction, malignant neoplasms of the trachea, bronchus, and lung, other chronic obstructive pulmonary disease, and all other forms of chronic ischemic heart disease.
+Our category of focus included filtering on illnesses like acute myocardial infarction, malignant neoplasms of the trachea, bronchus, and lung, other chronic obstructive pulmonary disease, and all other forms of chronic ischemic heart disease as seen in @fig-cause.
 Leading causes of death are measured and ranked by the top 30 most common death causes each specific year.
 The causes of death are classified based on the International Classification of Diseases 10th Edition.
-
-**AQHI Data**: The second dataset we used is the air quality health index (AQHI) dataset found at the government of Alberta’s open data portal @alberta.
-This dataset contains AQHI by municipality for the years 2012-2022 and reports air quality health index, and health risk both quantitatively and qualitatively.
-To use the AQHI dataset we employed simple data-cleaning practice to maintain descriptive variable names and readability.
-The data is measured by the percentage of hours for each year at a given air quality level, by municipality.
-The Air Quality Health Index is calculated based on the relative risks of a combination of common air pollutants that is known to harm human health.
-These pollutants are ozone (O3) at ground level, particulate matter (PM2.5), and nitrogen dioxide (NO2).
-Risks are defined as follows: 1-3 High Quality; 4-6 Moderate Quality; 7-9 Low Quality; 10+ Very Low Quality.
-
-**PM2.5 Data**: We used the PM2.5 data set retrieved from @alberta which was last updated in April 2023.
-It reports on average PM2.5 concentration levels through the years 2000-2021 using a provincial average, the 10th percentile quantities, and the 90th percentile quantities, with a focus on 8 municipalities Edmonton, Fort McMurray, Grande Prairie, Lethbridge, Medicine Hat, and Red Deer respectively and lastly reports the Canadian Ambient Air Quality Standard (CAAQS) value.
-
-The Alberta Air Zone report @report, which is linked to our dataset, provides a detailed explanation of the measurement and processing of the PM2.5 quantity.
-Alberta Air Zones divides Alberta into six air zones which are aligned with Alberta’s Land-use Framework regional boundaries.
-Ambient air quality in Alberta is monitored at continuous air monitoring stations located within these air zones.
-PM2.5 quantities are taken throughout these stations across Alberta, and they measure the quantities in µg (micrograms per cubic meter of air).
-
-## Data Cleaning
-
-We used R [@citeR] for data cleaning and processing, utilizing packages like tidyverse [@tidy] for data manipulation and janitor [@jan] for cleaning column names.
-Other packages used includes `ggplot2` [@ggplot], `dplyr` [@dp], `readr` [@read], `tibble` [@tib], `janitor` [@jan],`reshape2` [@reshape], `knitr` [@knit], `ggbeeswarm` [@gg], `ggrepel` [@repel], `kableExtra`[@kable], `readxl`[@readxl], `MASS`[@mass], `rstanarm`[@rstan], `modelsummary`[@model] and `here` [@here].
-
-The raw air quality data were preprocessed to remove inconsistencies and irrelevant information.
-Specifically, we filtered the dataset to include observations from the years 2012 to 2021, which are relevant to our analysis.
-Additionally, we merged this dataset with additional information on peak pollution levels for analysis.
-
-Similar to the previous datasets, the raw mortality data underwent cleaning procedures to focus on specific causes relevant to our analysis.
-We filtered the dataset to include observations up to 2021 and merged it with additional information on air quality for correlation analysis.
-
-The raw data on AQHI were filtered to include observations from the years 2011 to 2021 for consistency with other datasets.
-Additionally, the data were aggregated at the municipal level for further analysis.
-
-## Data Modifications
-
-In this study, we constructed unique datasets by thoughtfully selecting and merging data from the Government of Alberta's open data portal, Alberta.ca, spanning the years 2012 to 2022.
-Our process involved merging variables from various datasets to create specific datasets tailored for model building and analysis.
-One such dataset, 'cleaned_chart_data,' was created by merging variables such as causes of death, total deaths, provincial average PM2.5 levels, and CAAQS.
-This dataset was designed to facilitate our analysis of any significant correlations between these variables.
-A snapshot of this data is referenced in Table X.
-Additionally, we derived two other datasets, 'merged_data' and 'merged_heart_data,' by merging variables related to heart disease numbers, lung disease numbers, and provincial average PM2.5 values.
-These datasets were instrumental in examining the impact of PM2.5 on each type of illness, as previously discussed.
-Overall, our methodology ensured the creation of merged datasets that allowed for a detailed investigation into the relationships between PM2.5 levels and various health outcomes in Alberta.
-
-```{r}
-#| label: fig-lung
-#| fig-cap: Lung Related Causes Mortality Rates vs Average PM2.5 Quantity 
-#| echo: false
-#| warning: false
-
-# Plot the regression graph for all lungs causes
-ggplot(data_lung, aes(x = provincial_average, y = total_deaths)) +
-  geom_point() +
-  geom_smooth(method = "lm", se = FALSE) +  # Add linear regression line without confidence interval
-  labs(x = "Provincial Average", y = "Total Deaths") +
-  theme_minimal()
-```
-
-```{r}
-#| label: fig-heart
-#| fig-cap: Heart Related Causes Mortality Rates vs Average PM2.5 Quantity 
-#| echo: false
-#| warning: false
-# Plot the regression graph for all heart causes
-ggplot(data, aes(x = provincial_average, y = total_deaths)) +
-  geom_point() +
-  geom_smooth(method = "lm", se = FALSE) +  # Add linear regression line without confidence interval
-  labs(x = "Provincial Average", y = "Total Deaths") +
-  theme_minimal()
-
-```
-
-Talk more about it.
-
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
-
-```{r}
-#| label: fig-annual
-#| fig-cap: Annual Trends of PM2.5 Concentrations in Alberta
-#| echo: false
-#| warning: false
-#| message: false
-
-# Assuming your data frame is named 'cleaned_avg_PM25_conc' and is already filtered and renamed.
-# Convert factors to character to avoid issues in ggplot
-data[] <- lapply(data, function(x) if(is.factor(x)) as.character(x) else x)
-# Creating the plot with ggplot
-p <- ggplot(data, aes(x = year)) +
-  geom_line(aes(y = provincial_average, colour = "Provincial Average")) +
-  geom_line(aes(y = x10th_percentile, colour = "10th Percentile")) +
-  geom_line(aes(y = x90th_percentile, colour = "90th Percentile")) +
-  labs(x = "Years", y = "PM2.5 Concentration (μg/m³)") +
-  scale_colour_manual(values = c("Provincial Average" = "blue", "10th Percentile" = "green", "90th Percentile" = "red")) +
-  theme_minimal()
-
-# Display the plot
-p
-```
-
-```{r}
-#| label: fig-four
-#| fig-cap: Mortality Rates vs. Air Quality Health Index
-#| echo: false
-#| warning: false
-#| message: false
-
-library(MASS)
-
-air_quality_data <- air_quality_data[air_quality_data$air_quality_health_index != "10+",]
-air_quality_data$air_quality_health_index <- as.numeric(air_quality_data$air_quality_health_index)
-
-# Calculate the average Air Quality Health Index per year
-average_aqi_by_year <- air_quality_data %>%
-  group_by(year) %>%
-  summarize(average_aqi = mean(air_quality_health_index))
-
-combined_data <- merge(deaths_data, average_aqi_by_year, by = "year")
-
-# Plot scatter plot of mortality rates vs. air quality health index
-ggplot(combined_data, aes(x = average_aqi, y = total_deaths)) +
-  geom_point() +
-  geom_smooth(method = "lm") +
-  facet_wrap(~cause) +
-  labs(
-       x = "Average Air Quality Health Index",
-       y = "Total Deaths")
-```
 
@@
 
+**AQHI Data**: The second dataset we used is the air quality health index (AQHI) dataset found at the government of Alberta’s open data portal @alberta.
+This dataset contains AQHI by municipality for the years 2012-2022 and reports air quality health index, and health risk both quantitatively and qualitatively.
+To use the AQHI dataset we employed simple data-cleaning practice to maintain descriptive variable names and readability.
+The data is measured by the percentage of hours for each year at a given air quality level, by municipality.
+The Air Quality Health Index is calculated based on the relative risks of a combination of common air pollutants that is known to harm human health.
+These pollutants are ozone (O3) at ground level, particulate matter (PM2.5), and nitrogen dioxide (NO2).
+Risks are defined as follows: 1-3 High Quality; 4-6 Moderate Quality; 7-9 Low Quality; 10+ Very Low Quality.
+
+**PM2.5 Data**: We used the PM2.5 data set retrieved from @alberta which was last updated in April 2023.
+It reports on average PM2.5 concentration levels through the years 2000-2021 using a provincial average, the 10th percentile quantities, and the 90th percentile quantities, with a focus on 8 municipalities Edmonton, Fort McMurray, Grande Prairie, Lethbridge, Medicine Hat, and Red Deer respectively and lastly reports the Canadian Ambient Air Quality Standard (CAAQS) value.
+The trends of the PM2.5 concentration levels over the years can be seen in @fig-annual.
+
 ```{r}
-#| label: fig-average_air
-#| fig-cap: Average Air Quality Health Index by Year
+#| label: fig-annual
+#| fig-cap: Annual Trends of PM2.5 Concentrations in Alberta
 #| echo: false
@@
 #| message: false
-# Ensure the 'year' column is in numeric format for proper line plotting
-average_aqi_by_year$year <- as.numeric(as.character(average_aqi_by_year$year))
 
-# Plotting with ggplot2
-ggplot(average_aqi_by_year, aes(x = year, y = average_aqi)) +
-  geom_line() +  # for lines
-  geom_point() + # for points
-  labs(
-       x = "Year",
-       y = "Average Air Quality Index") +
-  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) # Optional: Adjust x-axis breaks
+# Assuming your data frame is named 'cleaned_avg_PM25_conc' and is already filtered and renamed.
+# Convert factors to character to avoid issues in ggplot
+data[] <- lapply(data, function(x) if(is.factor(x)) as.character(x) else x)
+# Creating the plot with ggplot
+p <- ggplot(data, aes(x = year)) +
+  geom_line(aes(y = provincial_average, colour = "Provincial Average")) +
+  geom_line(aes(y = x10th_percentile, colour = "10th Percentile")) +
+  geom_line(aes(y = x90th_percentile, colour = "90th Percentile")) +
+  labs(x = "Years", y = "PM2.5 Concentration (μg/m³)") +
+  scale_colour_manual(values = c("Provincial Average" = "blue", "10th Percentile" = "green", "90th Percentile" = "red")) +
+  theme_minimal()
+
+# Display the plot
+p
+```
+
+The Alberta Air Zone report @report, which is linked to our dataset, provides a detailed explanation of the measurement and processing of the PM2.5 quantity.
+Alberta Air Zones divides Alberta into six air zones which are aligned with Alberta’s Land-use Framework regional boundaries.
+Ambient air quality in Alberta is monitored at continuous air monitoring stations located within these air zones.
+PM2.5 quantities are taken throughout these stations across Alberta, and they measure the quantities in µg (micrograms per cubic meter of air).
+
+## Data Cleaning
+
+We used R [@citeR] for data cleaning and processing, utilizing packages like tidyverse [@tidy] for data manipulation and janitor [@jan] for cleaning column names.
+Other packages used includes `ggplot2` [@ggplot], `dplyr` [@dp], `readr` [@read], `tibble` [@tib], `janitor` [@jan],`reshape2` [@reshape], `knitr` [@knit], `ggbeeswarm` [@gg], `ggrepel` [@repel], `kableExtra`[@kable], `readxl`[@readxl], `MASS`[@mass], `rstanarm`[@rstan], `modelsummary`[@model] and `here` [@here].
+
+The raw air quality data were preprocessed to remove inconsistencies and irrelevant information as shown in @tbl-cleaned-data.
+Specifically, we filtered the dataset to include observations from the years 2012 to 2021, which are relevant to our analysis.
+Additionally, we merged this dataset with additional information on peak pollution levels for analysis.
+```{r}
+#| label: tbl-cleaned-data
+#| message: false
+#| echo: false
+#| warning: false
+#| tbl-cap: "Sample of Cleaned State Score Data"
+library(kableExtra)
+
+options(kableExtra.latex.max.width=1)
+head(air_quality_data, 6) |>
+  kable(
+        col.names = c("municipality","year","air_quality_health_index","health_risk","original_value"),
+    booktabs = TRUE,
+    longtable = FALSE
+  ) %>%
+  kable_styling(latex_options = c("striped", "scale_down"), font_size = 6)
+```
+Similar to the previous datasets, the raw mortality data(deaths-leading-causes data) underwent cleaning procedures to focus on specific causes relevant to our analysis.
+We filtered the dataset to include observations up to 2021 and merged it with additional information on air quality for correlation analysis as showed in @tbl-cleaned-data1.
+
+```{r}
+#| label: tbl-cleaned-data1
+#| message: false
+#| echo: false
+#| warning: false
+#| tbl-cap: "Sample of Cleaned State Score Data"
+library(kableExtra)
+
+options(kableExtra.latex.max.width=1)
+head(deaths_data, 6) |>
+  kable(
+        col.names = c("year","cause","ranking","total_deaths"),
+    booktabs = TRUE,
+    longtable = FALSE
+  ) %>%
+  kable_styling(latex_options = c("striped", "scale_down"), font_size = 6)
+```
+
+The PM2.5 Concentration data was cleaned and merged with other datasets. The Provincial Average of PM2.5 concentration per year is added it to the dataset of mortality to get better results.
+
+## Data Modifications
+
+In this study, we constructed unique datasets by thoughtfully selecting and merging data from the Government of Alberta's open data portal, Alberta.ca, spanning the years 2012 to 2022.
+Our process involved merging variables from various datasets to create specific datasets tailored for model building and analysis.
+One such dataset, 'cleaned_chart_data,' was created by merging variables such as causes of death, total deaths, provincial average PM2.5 levels, and CAAQS.
+This dataset was designed to facilitate our analysis of any significant correlations between these variables.
+
+Additionally, we derived two other datasets, 'merged_data' and 'merged_heart_data,' by merging variables related to heart disease numbers, lung disease numbers, and provincial average PM2.5 values from year 2001 to 2021.
+These datasets were instrumental in examining the impact of PM2.5 on each type of illness, as previously discussed.A snapshot of the 'merged_data" for the lung disease is referenced in @tbl-cleaned-data2.
+
+Overall, our methodology ensured the creation of merged datasets that allowed for a detailed investigation into the relationships between PM2.5 levels and various health outcomes in Alberta. 
+```{r}
+#| label: tbl-cleaned-data2
+#| message: false
+#| echo: false
+#| warning: false
+#| tbl-cap: "Sample of Cleaned State Score Data"
+library(kableExtra)
+
+options(kableExtra.latex.max.width=1)
+head(data_lung, 6) |>
+  kable(
+        col.names = c("year","total_deaths","provincial_average"),
+    booktabs = TRUE,
+    longtable = FALSE
+  ) %>%
+  kable_styling(latex_options = c("striped", "scale_down"), font_size = 6)
 ```
@@
 
-We used *Negative Binomial Regression (NBR)* to study how air pollution affects deaths related to breathing problems.
+We used **Negative Binomial Regression (NBR)** to study how air pollution affects deaths related to breathing problems.
 NBR is extremely useful for handling overdispersed count data, where the variance exceeds the mean @neg.
@@
 Negative Binomial Regression works better in these kinds of situations since it accounts for this extra unpredictability.
-It gives us easy-to-understand results that show how average air pollution in a province *(provincial_average)* can affect the number of deaths *(total_deaths)*.
+It gives us easy-to-understand results that show how average air pollution in a province **(provincial_average)** can affect the number of deaths **(total_deaths)**.
 
@@
 
-*Heart Diseases*:
+**Heart Diseases**:
 
@@
 
-*Lung Diseases*:
+**Lung Diseases**:
 
@@
 
-*Details of Regression Models*
+**Details of Regression Models**
 
@@
 
-1. *Modelling the count data*:
+1. **Modelling the count data**:
 
@@
 
-2. *Link function and predictors*
+2. **Link function and predictors**
 
@@
 
-3. *Coefficient*
+3. **Coefficient**
 
@@
 
-4. *Dispersion parameter*
+4. **Dispersion parameter**
 
@@
 
-*Model 1: Comparative Analysis of Mortality Causes*
+**Model 1: Comparative Analysis of Mortality Causes**
 
@@
 
-*Ischemic Heart Disease*: 
+**Ischemic Heart Disease**: 
 Both Poisson and negative binomial models showed a coefficient of 0.510, suggesting a consistent and significant association between air pollution and mortality due to Ischemic Heart Disease.
@@
 
-*Trachea/Bronchus/Lung Cancer*: Similarly, for Trachea/Bronchus/Lung Cancer, both models indicated a coefficient of 0.353, describing the impact of air pollution on these lung diseases.
+**Trachea/Bronchus/Lung Cancer**: Similarly, for Trachea/Bronchus/Lung Cancer, both models indicated a coefficient of 0.353, describing the impact of air pollution on these lung diseases.
 It is learned through this coefficient that increased levels of air pollution are associated with higher mortality rates from these types of lung cancer.
 
-*COPD*: For COPD, the coefficient was minimal (0.004), which suggests that, there is a negligible direct association between air pollution and COPD mortality.
+**COPD**: For COPD, the coefficient was minimal (0.004), which suggests that, there is a negligible direct association between air pollution and COPD mortality.
 It's important to note that this does not necessarily mean air pollution has no effect on COPD but rather, it might indicate that the relationship is more complex and possibly influenced by other variables not captured in the model.
@@
 ```
-*Model 2: Heart Disease and Air Quality*
+**Model 2: Heart Disease and Air Quality**
 
@@
 
-*Model 3: Lung Disease and Air Quality*
+**Model 3: Lung Disease and Air Quality**
 
@@
 
-Our results are summarized in @tbl-modelresults.
+## Correlation between AQHI and Causes/Death
+
+Exploring the relationship between AQHI and total mortality measured by the correlation graph @fig-four we notice there is a weak relationship between our chosen group of illnesses and AQHI.
+
+```{r}
+#| label: fig-four
+#| fig-cap: Mortality Rates vs. Air Quality Health Index
+#| echo: false
+#| warning: false
+#| message: false
+
+library(MASS)
+
+air_quality_data <- air_quality_data[air_quality_data$air_quality_health_index != "10+",]
+air_quality_data$air_quality_health_index <- as.numeric(air_quality_data$air_quality_health_index)
+
+# Calculate the average Air Quality Health Index per year
+average_aqi_by_year <- air_quality_data %>%
+  group_by(year) %>%
+  summarize(average_aqi = mean(air_quality_health_index))
+
+combined_data <- merge(deaths_data, average_aqi_by_year, by = "year")
+
+# Plot scatter plot of mortality rates vs. air quality health index
+ggplot(combined_data, aes(x = average_aqi, y = total_deaths)) +
+  geom_point() +
+  geom_smooth(method = "lm") +
+  facet_wrap(~cause) +
+  labs(
+       x = "Average Air Quality Health Index",
+       y = "Total Deaths")
+```
+
+The points representing mortality rates due to acute myocardial infarction seem relatively stable across the range of AQHI values.
+The fitted line is almost flat, suggesting that within the AQHI range provided, there is no strong relationship between AQHI and the mortality rates from this condition.
+
+Similar to acute myocardial infarction, the mortality rates for ‘All Other Forms of Chronic Ischemic Heart Disease’ appear stable across the AQHI values.
+The regression line is again flat, indicating no significant trend that suggests an increase or decrease in mortality rates associated with changes in AQHI within the observed range.
+
+For ‘Malignant Neoplasms of Trachea, Bronchus, and Lung’ the scatter plot shows a slight decrease in mortality rates with higher AQHI, as indicated by a gentle negative slope of the regression line.
+However, given the tight clustering of points, this trend might not be statistically significant.
+The data does not indicate a strong correlation between AQHI and mortality from these neoplasms within the given range.
+
+The trend for COPD appears similar to that of malignant neoplasms with a slight negative slope.
+However, the range of mortality rates is relatively narrow, and the points are closely clustered, suggesting a weak relationship, if any, between AQHI and mortality from COPD.
+
+While we observe a weak correlation between the AQHI and each type of illness, it is important to consider the trends in AQHI over the period from 2012 to 2022, as depicted in @fig-average_air.
+We notice a noticeable dip in AQHI around the COVID-19 years, but overall, the AQHI remains consistently below 5 on its 1-10 severity scale.
+This low peak in AQHI may not be severe enough to cause any correlation with the total mortalities of the respiratory and cardiac illnesses we are analyzing.
+Some of the weak relationships we assessed above can likely be attributed to this factor.
+
+```{r}
+#| label: fig-average_air
+#| fig-cap: Average Air Quality Health Index by Year
+#| echo: false
+#| warning: false
+#| message: false
+# Ensure the 'year' column is in numeric format for proper line plotting
+average_aqi_by_year$year <- as.numeric(as.character(average_aqi_by_year$year))
+
+# Plotting with ggplot2
+ggplot(average_aqi_by_year, aes(x = year, y = average_aqi)) +
+  geom_line() +  # for lines
+  geom_point() + # for points
+  labs(
+       x = "Year",
+       y = "Average Air Quality Index") +
+  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) # Optional: Adjust x-axis breaks
+```
+
+## Lung and Heart Related Mortality vs. PM2.5 Levels
+
+To assess how PM2.5 affected lung disease and heart disease we interpret the the trends by creating 2 different correlation plots.
+In the context of lung diseases, PM2.5 is expected to be a significant factor, whereas it should not have a significant impact on heart diseases.
+In @fig-lung, the lung-related causes graph, we observed a strong linear correlation between the provincial average of PM2.5 and the total deaths from lung diseases.
+As the provincial average of PM2.5 levels increased, so did the total death count from lung diseases.
+```{r}
+#| label: fig-lung
+#| fig-cap: Lung Related Causes Mortality Rates vs Average PM2.5 Quantity 
+#| echo: false
+#| warning: false
+
+# Plot the regression graph for all lungs causes
+ggplot(data_lung, aes(x = provincial_average, y = total_deaths)) +
+  geom_point() +
+  geom_smooth(method = "lm", se = FALSE) +  # Add linear regression line without confidence interval
+  labs(x = "Provincial Average", y = "Total Deaths") +
+  theme_minimal()
+```
+
+In contrast, the heart disease-related deaths showed only a slight correlation when compared to the same levels of PM2.5, as seen in @fig-heart .
+The values of PM2.5 did not appear to significantly affect heart-related diseases, as the increase in PM2.5 levels resulted in a weak positive linear relationship with the total heart disease-related deaths.
+
+```{r}
+#| label: fig-heart
+#| fig-cap: Heart Related Causes Mortality Rates vs Average PM2.5 Quantity 
+#| echo: false
+#| warning: false
+# Plot the regression graph for all heart causes
+ggplot(data, aes(x = provincial_average, y = total_deaths)) +
+  geom_point() +
+  geom_smooth(method = "lm", se = FALSE) +  # Add linear regression line without confidence interval
+  labs(x = "Provincial Average", y = "Total Deaths") +
+  theme_minimal()
+
+```
 
@@
 
-## First discussion point {#sec-first-point}
+## Findings
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+Combining the insights from our models and the graphical data from the graphs for lung and heart-related mortality rates versus average PM2.5 levels, we can conclude the following overall findings from our analysis:
+ AQHI did not have a significant correlation with respiratory and cardiac illness mortality.
+PM2.5 had a significant association with lung-related illness mortality.
+PM2.5 had a weak correlation to heart-related diseases and, overall did not correlate to its mortality numbers.
 
-## Second discussion point
+More specifically we discuss the model findings, possible reasons for our results, and discrepancies in our correlation analysis and model results below.
 
-## Third discussion point
+The PM2.5 regression models demonstrate that there is a statistically significant positive association between PM2.5 levels and mortality rates for Ischemic Heart Disease and Trachea/Bronchus/Lung Cancer, as indicated by positive coefficients.
+For COPD and other heart-related causes, the models suggested no direct association with air pollution within the data scope that was analyzed.
 
-## Weaknesses and next steps
+The correlation graph for lung-related causes and PM2.5 levels shows a slight upward trend, indicating a potential increase in mortality rates with higher PM2.5 levels.
+This corresponds with the positive coefficients found in the regression models, supporting the hypothesis that poor air quality can affect lung health.
 
-Weaknesses and next steps should also be included.
+However, the graph for heart-related causes shows a relatively flat trend, which suggests a weak relationship between PM2.5 levels and mortality rates.
+However, the regression model indicated a significant relationship for Ischemic Heart Disease, overall these results likely suggest that other factors could be influencing the mortality rates that are not captured in this analysis.
+
+The differences in results between the model analysis and correlation graphs can be attributed to several factors.
+Firstly, the complexity of environmental health relationships means that simple linear representations may not capture all relationships without the benefit of controlling for other variables.
+The regression model's focus on specific parameters may also lead to discrepancies, as it may not fully account for all the variables at play.
+The nature of the data and the limitations of the models themselves can contribute to discrepancies between the two analyses.
+Additionally, the models account for overdispersion and other confounding variables, providing a more informed understanding of the relationship between air pollution and mortality rates than the data visualizations alone.
+
+Overall, the findings from the models, supported by the trends from the graphs, suggest that PM2.5 has a varying impact on different health outcomes.
+There is a clear indication that higher levels of PM2.5 are associated with increased mortality from lung cancer and Ischemic Heart Disease.
+Alternatively, the weak graphical correlation in the heart-related graph does not diminish the relevance of the model's findings but rather underscores the need for a thorough analysis that accounts for a variety of factors affecting heart health.
+
+## Bias
+
+Our analysis is based on the datasets available from provincial open databases, potentially introducing selection bias as these datasets may not represent the entire population of Alberta accurately.
+Certain demographics or regions may be omitted from the analysis, leading to biased conclusions.
+Our study does not account for all potential variables that could influence the relationship between air quality and mortality rates, such as socioeconomic factors, healthcare access, lifestyle habits, and weather conditions.
+Since we are only analyzing a few variables to find correlations towards respiratory-related mortality, these variables could introduce bias into the analysis.
+The datasets used in the analysis may also suffer from sampling bias, particularly if they do not adequately capture the entire population or if certain years or regions are overrepresented compared to others.
+This could affect the reliability of the findings.
+As a result, the accuracy and integrity of the data heavily depend on the government of Alberta’s data collection reporting and practice for the three datasets used in our study.
+
+## Limitations
+
+In our study, there might be several limitations affecting our conclusions.
+First, between 2010 and 2017, older equipment at Alberta’s monitoring stations was replaced with new monitoring equipment for PM2.5.
+These new instruments measured an additional portion (semi-volatile) of the PM2.5 mass not captured by older instruments.
+As a result, the older monitoring equipment used between 2000 and 2017 likely underreported concentrations of PM2.5 under some conditions, after 2010, the increase in PM2.5 concentrations may be a result of changes in monitoring equipment (PM2.5 reference).
+Overall, the concentrations measured with the new monitors may not be directly comparable with measurements from years in which older instruments were used, possibly hindering trends in our study by some amount.
+
+Furthermore, our study's analysis is limited by the time frame of the available datasets, which may not capture long-term trends or changes in air quality and mortality rates over extended periods.
+As a result, our findings may not be valid if generalized to other regions with different environmental and demographic characteristics.
+While the study aims to investigate the relationship between air quality and mortality rates, it may be challenging to establish causation based solely on observational data.
+The findings may indicate correlations between air quality and mortality rates but cannot definitively prove causality.
+
+Lastly, our datasets used are of relatively small size, and this may have significantly impacted the statistical power of the analysis.
+With fewer data points, the ability to detect meaningful relationships or patterns between variables may have deterred our models from finding any underlying correlations that were not found with the current size.
+This limitation may have reduced the reliability of the findings we concluded.
 
@@
 
-# Additional data details
-
 # Model details {#sec-model-details}
@@
 
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...
+Posterior Predictive Checks for General Model (Death Causes and Total Deaths):
 
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...
+In @fig-ppcheckfirst-1 (Posterior prediction check): This plot indicates how well the model's predictions align with the observed data.
+The overlap between the observed counts (y) and the posterior predictions (y_rep) suggests the model captures the underlying distribution of total death causes effectively.
+
+In @fig-ppcheckfirst-2 (Comparing the posterior with the prior): The plot displays a refinement from broad prior beliefs to more precise posterior beliefs after incorporating the data.
+The tighter posterior distribution signals that the data have had a substantial impact on the model's estimates.
+
+Posterior Predictive Checks for Heart and Lung Models:
+
+In @fig-ppcheckheartlung-1 (Posterior prediction check for the heart model): The graph shows the posterior predictions closely mirroring the observed mortality data, indicating that the heart model accurately reflects the real-world distribution of heart-related deaths.
+
+In @fig-ppcheckheartlung-2 (Posterior prediction check for the lung model): Similar to the heart model, this plot demonstrates that the lung model's predictions are consistent with the observed data, reaffirming the model's validity in describing lung-related mortality patterns.
 
@@
 ## Diagnostics
+@fig-stanareyouokay-1: This trace plot shows the sampled values over iterations for various model parameters.
+The chains appear stable and well-mixed around a consistent mean, suggesting good convergence of the MCMC algorithm.
+@fig-stanareyouokay-3: The trace plot for the lung model indicates that the chains for each parameter exhibit consistent behavior across iterations, without trending upwards or downwards, which implies convergence.
+@fig-stanareyouokay-5: For the heart model, this trace plot displays a similar pattern of stability and suggests that the chains have converged, as evidenced by their hovering around a constant value.
+Rhat Plots:
 
-@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...
-
-@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...
+@fig-stanareyouokay-2: This Rhat plot displays values close to 1.00 for each chain, indicating that convergence has likely been achieved for the first model's parameters.
+@fig-stanareyouokay-4: The Rhat values for the lung model are near the ideal of 1.00, suggesting that there is no evidence of non-convergence and that the model is well-fitted.
+@fig-stanareyouokay-6: In the heart model's Rhat plot, the Rhat values are again close to 1.00, indicating successful convergence across the chains for all estimated parameters.
 

### 2024-03-16 paper/datasheet.qmd
Last minute cleanups

---
title: "Datasheet For Air Quality Index By Municipality, Leading Causes Of Death And Average PM2.5 Concentrated Dataset"
author: 
  - Vanshika Vanshika
  - Shivank Goel
  - Navya Hooda
thanks: "Code and data are available at: https://github.com/shivankgoel003/Mortality-in-Alberta."
date: today
date-format: long
abstract: "This datasheet documents the datasets used in a study analyzing the impact of air pollution on health outcomes in Alberta, Canada. The data for the study is sourced from the Alberta government's open data portal. It spans several years and includes variables like AQHI readings, PM2.5 concentrations, and mortality rates linked to specific health conditions. This dataset is helpful for its concise representation to study the environmental health in Alberta. It offers insights into public health implications of air pollution. The data cleaning and processing done using tools like R, ensures its reliability and accuracy. This datasheet is an essential guide to the dataset's structure, collection process, and potential applications."

format: pdf
number-sections: true
bibliography: references.bib
---

Extract of the questions from @gebru2021datasheets.

**Motivation**

1. *For what purpose was the dataset created? Was there a specific task in mind? Was there a specific gap that needed to be filled? Please provide a description.*

The dataset was created to investigate the correlation between air quality, particularly PM2.5 levels, and mortality rates due to respiratory and cardiac illnesses in Alberta. The aim was to understand how air pollution impacts public health in this region.

2. *Who created the dataset (for example, which team, research group) and on behalf of which entity (for example, company, institution, organization)?*

The datasets were created by the Alberta Government, specifically by Jobs, Economy and Northern Development, Service Alberta, and Environment and Parks departments.

3. *Who funded the creation of the dataset? If there is an associated grant, please provide the name of the grantor and the grant name and number.*

The creation of these datasets was likely funded by the Alberta government. No specific grant details were provided.

4. *Any other comments?*

These datasets provide critical insights into environmental and public health concerns in Alberta and are vital for ongoing research and policy development in these areas.

**Composition**

1. *What do the instances that comprise the dataset represent (for example, documents, photos, people, countries)? Are there multiple types of instances (for example, movies, users, and ratings; people and interactions between them; nodes and edges)? Please provide a description.*

The instances represent various types of environmental and public health data. The first dataset details the Air Quality Health Indices by municipality in Alberta. The second dataset provides information on the leading causes of death in Alberta. The third dataset offers data on PM2.5 concentrations in Alberta.

2. *How many instances are there in total (of each type, if appropriate)?*

The exact number of instances is not specified, but each dataset covers multiple years of data across various locations in Alberta.

3. *Does the dataset contain all possible instances or is it a sample (not necessarily random) of instances from a larger set? If the dataset is a sample, then what is the larger set? Is the sample representative of the larger set (for example, geographic coverage)? If so, please describe how this representativeness was validated/verified. If it is not representative of the larger set, please describe why not (for example, to cover a more diverse range of instances, because instances were withheld or unavailable).*

The datasets seem to be comprehensive for their respective scopes, covering broadly geographical areas within Alberta. It seems to be representative of the larger set of environmental and health data within the province.

4. *What data does each instance consist of? "Raw" data (for example, unprocessed text or images) or features? In either case, please provide a description.*

Each dataset has specific labels or targets. For the air quality index, the labels include AQHI values and municipal regions. The causes of death dataset labels include types of illnesses and mortality counts. The PM2.5 dataset's labels consist of concentration levels, locations, and time periods

5. *Is there a label or target associated with each instance? If so, please provide a description.*

Each dataset has specific labels or targets. For the air quality index, the labels include AQHI values and municipal regions. The causes of death dataset labels include types of illnesses and mortality counts. The PM2.5 dataset's labels consist of concentration levels, locations, and time periods.

6. *Is any information missing from individual instances? If so, please provide a description, explaining why this information is missing (for example, because it was unavailable). This does not include intentionally removed information, but might include, for example, redacted text.*

There is no specific mention of missing information. However, as with most large datasets, there may be gaps due to data unavailability or limitations in data collection.

7. *Are relationships between individual instances made explicit (for example, users' movie ratings, social network links)? If so, please describe how these relationships are made explicit.*

The datasets do not explicitly mention relationships between individual instances, but they can be inferred. For example, higher AQHI and PM2.5 values may correlate with increased mortality rates from certain diseases.

8. *Are there recommended data splits (for example, training, development/validation, testing)? If so, please provide a description of these splits, explaining the rationale behind them.*

There are no specific recommended data splits mentioned. Data splits would depend on the specific research or analysis goals.

9. *Are there any errors, sources of noise, or redundancies in the dataset? If so, please provide a description.*

Specific errors or sources of noise are not mentioned. However, as with any environmental data, there may be variabilities and inconsistencies due to natural fluctuations and measurement limitations.

10. *Is the dataset self-contained, or does it link to or otherwise rely on external resources (for example, websites, tweets, other datasets)? If it links to or relies on external resources, a) are there guarantees that they will exist, and remain constant, over time; b) are there official archival versions of the complete dataset (that is, including the external resources as they existed at the time the dataset was created); c) are there any restrictions (for example, licenses, fees) associated with any of the external resources that might apply to a dataset consumer? Please provide descriptions of all external resources and any restrictions associated with them, as well as links or other access points, as appropriate.*

The datasets are self-contained and available through the Alberta open data portal. They are subject to the Open Government Licence - Alberta, which allows for reuse and distribution.

11. *Does the dataset contain data that might be considered confidential (for example, data that is protected by legal privilege or by doctor-patient confidentiality, data that includes the content of individuals' non-public communications)? If so, please provide a description.*

No confidential or personally identifiable information is included in the dataset.

12. *Does the dataset contain data that, if viewed directly, might be offensive, insulting, threatening, or might otherwise cause anxiety? If so, please describe why.*

There is no content in these datasets that would be considered offensive or insulting. The data is factual, focusing on environmental and health statistics.

13. *Does the dataset identify any sub-populations (for example, by age, gender)? If so, please describe how these subpopulations are identified and provide a description of their respective distributions within the dataset.*

The datasets do not specify sub-populations like age or gender. They are more focused on geographical and temporal aspects of air quality and health outcomes.

14. *Is it possible to identify individuals (that is, one or more natural persons), either directly or indirectly (that is, in combination with other data) from the dataset? If so, please describe how.*

Individual identification is not possible with these datasets as they deal with aggregated data at the municipal level or higher.

15. *Does the dataset contain data that might be considered sensitive in any way (for example, data that reveals race or ethnic origins, sexual orientations, religious beliefs, political opinions or union memberships, or locations; financial or health data; biometric or genetic data; forms of government identification, such as social security numbers; criminal history)? If so, please provide a description.*

The datasets do not contain sensitive personal data. They are focused on environmental measures and aggregated health outcomes.

16. *Any other comments?*
	N/A

**Collection process**

1. *How was the data associated with each instance acquired? Was the data directly observable (for example, raw text, movie ratings), reported by subjects (for example, survey responses), or indirectly inferred/derived from other data (for example, part-of-speech tags, model-based guesses for age or language)? If the data was reported by subjects or indirectly inferred/derived from other data, was the data validated/verified? If so, please describe how.*

The data were directly observed and collected through environmental monitoring stations and health records. They are likely verified through standard government data validation processes, though specific details are not provided.

2. *What mechanisms or procedures were used to collect the data (for example, hardware apparatuses or sensors, manual human curation, software programs, software APIs)? How were these mechanisms or procedures validated?*

The air quality and PM2.5 data were likely collected using environmental monitoring stations equipped with sensors. The causes of death data were probably compiled from health records.

3. *If the dataset is a sample from a larger set, what was the sampling strategy (for example, deterministic, probabilistic with specific sampling probabilities)?*

The datasets seem to be comprehensive collections rather than samples. They likely  represent the entire population of interest (i.e., all municipalities in Alberta for air quality and all recorded causes of death).

4. *Who was involved in the data collection process (for example, students, crowdworkers, contractors) and how were they compensated (for example, how much were crowdworkers paid)?*

Data collection was likely conducted by government employees or contractors. Specific details about their compensation are not provided.

5. *Over what timeframe was the data collected? Does this timeframe match the creation timeframe of the data associated with the instances (for example, recent crawl of old news articles)? If not, please describe the timeframe in which the data associated with the instances was created.*

The data collection covers several years, as indicated by the datasets (e.g., 2001-2021 for the leading causes of death).

6. *Were any ethical review processes conducted (for example, by an institutional review board)? If so, please provide a description of these review processes, including the outcomes, as well as a link or other access point to any supporting documentation.*

There is no specific mention of ethical review processes.

7. *Did you collect the data from the individuals in question directly, or obtain it via third parties or other sources (for example, websites)?*

The data were collected from environmental monitoring stations and health records, not directly from individuals. It is a secondary data source.

8. *Were the individuals in question notified about the data collection? If so, please describe (or show with screenshots or other information) how notice was provided, and provide a link or other access point to, or otherwise reproduce, the exact language of the notification itself.*

Individual notification is not applicable, as the data do not pertain to individual persons.

9. *Did the individuals in question consent to the collection and use of their data? If so, please describe (or show with screenshots or other information) how consent was requested and provided, and provide a link or other access point to, or otherwise reproduce, the exact language to which the individuals consented.*

Individual consent is not relevant for this type of data, as it does not involve personal data collection from individuals.

10. *If consent was obtained, were the consenting individuals provided with a mechanism to revoke their consent in the future or for certain uses? If so, please provide a description, as well as a link or other access point to the mechanism (if appropriate).*

As individual consent was not required for this data collection.

11. *Has an analysis of the potential impact of the dataset and its use on data subjects (for example, a data protection impact analysis) been conducted? If so, please provide a description of this analysis, including the outcomes, as well as a link or other access point to any supporting documentation.*

There is no information provided regarding an impact analysis on data subjects, likely because the datasets deal with  environmental and health data and not individual-specific data.

12. *Any other comments?*

The data collection process is extensive and critical for understanding the environmental and public health in Alberta. The absence of personal data reduces privacy concerns.

**Preprocessing/cleaning/labeling**

1. *Was any preprocessing/cleaning/labeling of the data done (for example, discretization or bucketing, tokenization, part-of-speech tagging, SIFT feature extraction, removal of instances, processing of missing values)? If so, please provide a description. If not, you may skip the remaining questions in this section.*

There is likely some level of preprocessing and cleaning involved, ensuring consistency, and possibly dealing with missing values. However, specific details are not provided in the information given.

2. *Was the "raw" data saved in addition to the preprocessed/cleaned/labeled data (for example, to support unanticipated future uses)? If so, please provide a link or other access point to the "raw" data.*

There is no mention of whether the raw data is saved alongside the processed data.

3. *Is the software that was used to preprocess/clean/label the data available? If so, please provide a link or other access point.*

Details about the specific software or tools used for preprocessing, cleaning, or labeling are not provided.

4. *Any other comments?*

Proper data preprocessing and cleaning are essential for the accuracy and reliability of analysis results. The datasets, as released, should be ready for use in various analyses and research.

**Uses**

1. *Has the dataset been used for any tasks already? If so, please provide a description.*

The datasets have likely been used for academic research and study before. They have been used to analyze the relationship between mortality rates and causes of certain diseases.

2. *Is there a repository that links to any or all papers or systems that use the dataset? If so, please provide a link or other access point.*

No specific repository is mentioned, but the datasets are available through the Alberta open data portal.

3. *What (other) tasks could the dataset be used for?*

The data could be used for further environmental health studies, public health planning, and educational purposes.

4. *Is there anything about the composition of the dataset or the way it was collected and preprocessed/cleaned/labeled that might impact future uses? For example, is there anything that a dataset consumer might need to know to avoid uses that could result in unfair treatment of individuals or groups (for example, stereotyping, quality of service issues) or other risks or harms (for example, legal risks, financial harms)? If so, please provide a description. Is there anything a dataset consumer could do to mitigate these risks or harms?*

The users should consider potential biases and reporting in data collection.Also they should be aware of geographic
coverage to avoid misinterpretations.

5. *Are there tasks for which the dataset should not be used? If so, please provide a description.*

The data should  not be used for identifying or targeting specific individuals.

6. *Any other comments?*

The datasets are valuable for understanding trends and correlations in environmental and public health. They should be used appropriately, considering their scope and limitations.

**Distribution**

1. *Will the dataset be distributed to third parties outside of the entity (for example, company, institution, organization) on behalf of which the dataset was created? If so, please provide a description.*

The datasets are publicly available and can be accessed by third parties. They are distributed under the Open Government Licence - Alberta, which allows for use and distribution by anyone.

2. *How will the dataset be distributed (for example, tarball on website, API, GitHub)? Does the dataset have a digital object identifier (DOI)?*

The datasets are available for download through the Alberta open data portal. There is no mention of a DOI or distribution through other platforms like GitHub.

3. *When will the dataset be distributed?*

The datasets are already available for public access.

4. *Will the dataset be distributed under a copyright or other intellectual property (IP) license, and/or under applicable terms of use (ToU)? If so, please describe this license and/ or ToU, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms or ToU, as well as any fees associated with these restrictions.*

The datasets are distributed under the Open Government Licence - Alberta, which permits a wide range of uses. The license details can be found on the Alberta open data portal.

5. *Have any third parties imposed IP-based or other restrictions on the data associated with the instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms, as well as any fees associated with these restrictions.*

There is no mention of any third-party IP restrictions on these datasets.

6. *Do any export controls or other regulatory restrictions apply to the dataset or to individual instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any supporting documentation.*

There are no specified export controls or regulatory restrictions beyond the Open Government Licence - Alberta.

7. *Any other comments?*

The open access nature of these datasets makes them a valuable resource for a wide range of users, from researchers to policymakers.

**Maintenance**

1. *Who will be supporting/hosting/maintaining the dataset?*

The datasets are maintained by the respective departments of the Alberta government – Jobs, Economy and Northern Development, Service Alberta, and Environment and Parks.

2. *How can the owner/curator/manager of the dataset be contacted (for example, email address)?*

Contact information is likely available through the Alberta government’s open data portal.

3. *Is there an erratum? If so, please provide a link or other access point.*

There is no mention of an erratum.

4. *Will the dataset be updated (for example, to correct labeling errors, add new instances, delete instances)? If so, please describe how often, by whom, and how updates will be communicated to dataset consumers (for example, mailing list, GitHub)?*

The datasets are updated annually. Specific communication channels for updates are not mentioned.

5. *If the dataset relates to people, are there applicable limits on the retention of the data associated with the instances (for example, were the individuals in question told that their data would be retained for a fixed period of time and then deleted)? If so, please describe these limits and explain how they will be enforced.*

There is no specific information about the maintenance of older versions of the dataset.

6. *Will older versions of the dataset continue to be supported/hosted/maintained? If so, please describe how. If not, please describe how its obsolescence will be communicated to dataset consumers.*

This is not applicable as the datasets do not contain individual-level data.

7. *If others want to extend/augment/build on/contribute to the dataset, is there a mechanism for them to do so? If so, please provide a description. Will these contributions be validated/verified? If so, please describe how. If not, why not? Is there a process for communicating/distributing these contributions to dataset consumers? If so, please provide a description.*

There is no mentioned mechanism for external contributions to the dataset. The data are sourced from government monitoring and records, so external contributions might not be applicable.

8. *Any other comments?*

Regular maintenance and updates are crucial for the continued relevance and accuracy of these datasets.

\newpage

# References

### 2024-03-16 paper/paper.qmd
Final Paper

@@
 date-format: long
-abstract: "Air pollution's severe health impact on human respiratory and cardiac systems is a growing public health issue.
-The fine particulate matter (PM2.5) arise a primary concern due to its deep lung penetration and systemic effects.
-This research uses data from Alberta, Canada, to study the connection between PM2.5 levels as measured by the Air Quality Health Index (AQHI),  and the number of deaths from certain heart and lung diseases.
-We utilized statistical models, and identified a strong association between increased PM2.5 levels and higher mortality from these diseases.
-These findings suggest that there is an immediate need for health policies that address air quality improvements to safeguard public health for Alberta and beyond."
+abstract: "Air pollution's severe impact on human respiratory and cardiac health is a growing public health concern across the world.
+The fine particulate matter (PM2.5) air pollutant is of particular concern because it can deeply penetrate the lungs and cause systemic effects and is largely found in urban environments.
+This research uses data from Alberta, Canada to assess any correlations between PM2.5 levels and the regional Air Quality Health Index (AQHI), against the number of deaths from groups of respiratory and cardiac illnesses we selected.
+We utilized statistical models and identified a strong association between increased PM2.5 levels and higher mortality from these diseases.
+These findings suggest that there is an immediate need for health policies that address air quality improvements to safeguard public health in Alberta and beyond."
 format: pdf
@@
 data <- read.csv(here("data/analysis_data/cleaned_chart_data.csv"))
-data_peak <- read.csv(here("data/analysis_data/cleaned_peak_data.csv"))
+#data_peak <- read.csv(here("data/analysis_data/cleaned_peak_data.csv"))
 data_lung <- read.csv(here("data/analysis_data/merged_data.csv"))
@@
 neg_binom_model <- readRDS(file = here::here("models/neg_binom_model.rds"))
-stan_neg_binom_model <- readRDS(file = here::here("models/stan_neg_binom_model.rds"))
-stan_poisson_model <-readRDS(file = here::here("models/stan_poisson_model.rds"))
+#stan_neg_binom_model <- readRDS(file = here::here("models/stan_neg_binom_model.rds"))
+#stan_poisson_model <-readRDS(file = here::here("models/stan_poisson_model.rds"))
 
@@
 
+## Future Research
+
+While this study provides valuable insights into the relationship between air quality and mortality rates in Alberta, further research is needed to comprehensively understand the underlying mechanisms and potential confounding factors influencing this relationship.
+Future studies could explore additional variables, such as socioeconomic status, lifestyle factors, and healthcare access, to elucidate the complex interplay between air quality and mortality rates.
+Additionally,  studies spanning longer periods  could provide insights into the cumulative effects of chronic exposure to poor air quality on population health.
+Such research would be useful for informing evidence-based public health interventions and policy decisions aimed at promoting clean air and protecting human health.
+
 \newpage

### 2024-03-16 paper/paper.qmd
FINAL PAPER

@@
 abstract: "Air pollution's severe impact on human respiratory and cardiac health is a growing public health concern across the world.
-The fine particulate matter (PM2.5) air pollutant is of particular concern because it can deeply penetrate the lungs and cause systemic effects and is largely found in urban environments.
+The fine particulate matter (PM2.5) air pollutant is of particular concern because it can deeply penetrate the lungs and is largely found in urban environments.
 This research uses data from Alberta, Canada to assess any correlations between PM2.5 levels and the regional Air Quality Health Index (AQHI), against the number of deaths from groups of respiratory and cardiac illnesses we selected.
@@
 These new instruments measured an additional portion (semi-volatile) of the PM2.5 mass not captured by older instruments.
-As a result, the older monitoring equipment used between 2000 and 2017 likely underreported concentrations of PM2.5 under some conditions, after 2010, the increase in PM2.5 concentrations may be a result of changes in monitoring equipment (PM2.5 reference).
+As a result, the older monitoring equipment used between 2000 and 2017 likely underreported concentrations of PM2.5 under some conditions, after 2010, the increase in PM2.5 concentrations may be a result of changes in monitoring equipment @pm.
 Overall, the concentrations measured with the new monitors may not be directly comparable with measurements from years in which older instruments were used, possibly hindering trends in our study by some amount.
